---
title: "Propensity Score Weighting"
author: "Lucy D'Agostino McGowan"
institute: "Wake Forest University"
date: "2020-07-29 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center", dpi = 320, fig.height = 4)
```

class: inverse

## Propensity scores

* Weighting
* Matching
* Stratification
* Direct Adjustment
* ...

---
class: inverse

## Propensity scores

* **Weighting**
* Matching
* Stratification
* Direct Adjustment
* ...

---
class: inverse

## Target estimands

### Average Treatment Effect (ATE)

$$\Large w_{ATE} = \frac{Z_i}{p_i} + \frac{1-Z_i}{1 - p_i}$$

---
class: inverse

## Target estimands

### Average Treatment Effect Among the Treated (ATT)
$$\Large w_{ATT} = \frac{p_i Z_i}{p_i} + \frac{p_i (1-Z_i)}{1-p_i}$$
--

### Average Treatment Effect Among the Controls (ATC)
$$\Large w_{ATC} = \frac{(1-p_i)Z_i}{p_i} + \frac{(1-p_i)(1-Z_i)}{(1-p_i)}$$

---
class: inverse

## Target estimands

### Average Treatment Effect Among the Evenly Matchable (ATM)
$$\Large w_{ATM} = \frac{\min \{p_i, 1-p_i\}}{z_ip_i + (1-Z_i)(1-p_i)}$$
--

### Average Treatment Effect Among the Overlap Population
$$\Large w_{ATO} = (1-p_i)Z_i + p_i(1-Z_i)$$

---

```{r, include = FALSE}
library(tidyverse)
library(broom)
library(cidata)
propensity_model <- glm(
  qsmk ~ sex + 
    race + age + I(age^2) + education + 
    smokeintensity + I(smokeintensity^2) + 
    smokeyrs + I(smokeyrs^2) + exercise + active + 
    wt71 + I(wt71^2), 
  family = binomial(), 
  data = nhefs_complete
)

df <- propensity_model %>% 
  augment(type.predict = "response", data = nhefs_complete) %>% 
  mutate(wts = 1 / ifelse(qsmk == 0, 1 - .fitted, .fitted),
    w_ate = (qsmk / .fitted) + 
      ((1 - qsmk) / (1 - .fitted)),
    w_att = ((.fitted * qsmk) / .fitted) + 
      ((.fitted * (1 - qsmk)) / (1 - .fitted)),
    w_atc = (((1 - .fitted) * qsmk) / .fitted) + 
      (((1 - .fitted) * (1 - qsmk)) / (1 - .fitted)),
    w_atm = pmin(.fitted, 1 - .fitted) / 
      (qsmk * .fitted + (1 - qsmk) * (1 - .fitted)),
    w_ato = (1 - .fitted) * qsmk + 
      .fitted * (1 - qsmk)
  )

d <- df %>%
  tidyr::spread(qsmk, .fitted, sep = "_p")
```

## Histogram of propensity scores

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(d) + 
  geom_histogram(bins = 50, aes(qsmk_p1)) + 
  geom_histogram(bins = 50, aes(x = qsmk_p0, y = -..count..)) + 
  ylab("count") + xlab("p") +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs) 
```

---

## ATE

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(d) +
  geom_histogram(bins = 50, aes(qsmk_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(qsmk_p1, weight = w_ate), fill = "green", alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = qsmk_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = qsmk_p0, weight = w_ate, y = -..count..), fill = "blue", alpha = 0.5) + 
  ylab("count") + xlab("p") +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs) +
    theme_minimal() + 
  geom_rect(aes(xmin = 0.95, xmax = 1, ymin = 5, ymax = 100), fill = "#5DB854") + 
  geom_text(aes(x = 0.975, y = 50), label = "trt", angle = 270, color = "white") + 
   geom_rect(aes(xmin = 0.95, xmax = 1, ymin = -100, ymax = -5), fill = "#5154B8") + 
  geom_text(aes(x = 0.975, y = -50), label = "control", angle = 270, color = "white")
```

---

## ATT


```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(d) +
  geom_histogram(bins = 50, aes(qsmk_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(qsmk_p1, weight = w_att), fill = "green", alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = qsmk_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = qsmk_p0, weight = w_att, y = -..count..), fill = "blue", alpha = 0.5) + 
  ylab("count") + xlab("p") +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs) +
    theme_minimal() + 
  geom_rect(aes(xmin = 0.95, xmax = 1, ymin = 5, ymax = 30), fill = "#5DB854") + 
  geom_text(aes(x = 0.975, y = 17), label = "trt", angle = 270, color = "white") + 
   geom_rect(aes(xmin = 0.95, xmax = 1, ymin = -100, ymax = -5), fill = "#5154B8") + 
  geom_text(aes(x = 0.975, y = -50), label = "control", angle = 270, color = "white")
```


---

## ATC

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(d) +
  geom_histogram(bins = 50, aes(qsmk_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(qsmk_p1, weight = w_atc), fill = "green", alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = qsmk_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = qsmk_p0, weight = w_atc, y = -..count..), fill = "blue", alpha = 0.5) + 
  ylab("count") + xlab("p") +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs) +
    theme_minimal() + 
  geom_rect(aes(xmin = 0.95, xmax = 1, ymin = 5, ymax = 100), fill = "#5DB854") + 
  geom_text(aes(x = 0.975, y = 50), label = "trt", angle = 270, color = "white") + 
   geom_rect(aes(xmin = 0.95, xmax = 1, ymin = -100, ymax = -5), fill = "#5154B8") + 
  geom_text(aes(x = 0.975, y = -50), label = "control", angle = 270, color = "white")
```

---

## ATM

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(d) +
  geom_histogram(bins = 50, aes(qsmk_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(qsmk_p1, weight = w_atm), fill = "green", alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = qsmk_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = qsmk_p0, weight = w_atm, y = -..count..), fill = "blue", alpha = 0.5) + 
  ylab("count") + xlab("p") +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs) +  theme_minimal() + 
  geom_rect(aes(xmin = 0.95, xmax = 1, ymin = 5, ymax = 30), fill = "#5DB854") + 
  geom_text(aes(x = 0.975, y = 17), label = "trt", angle = 270, color = "white") + 
   geom_rect(aes(xmin = 0.95, xmax = 1, ymin = -100, ymax = -5), fill = "#5154B8") + 
  geom_text(aes(x = 0.975, y = -50), label = "control", angle = 270, color = "white")
```

---

## ATO

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(d) +
  geom_histogram(bins = 50, aes(qsmk_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(qsmk_p1, weight = w_ato), fill = "green", alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = qsmk_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = qsmk_p0, weight = w_ato, y = -..count..), fill = "blue", alpha = 0.5) + 
  ylab("count") + xlab("p") +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs) +
  theme_minimal() + 
  geom_rect(aes(xmin = 0.95, xmax = 1, ymin = 5, ymax = 30), fill = "#5DB854") + 
  geom_text(aes(x = 0.975, y = 17), label = "trt", angle = 270, color = "white") + 
   geom_rect(aes(xmin = 0.95, xmax = 1, ymin = -100, ymax = -5), fill = "#5154B8") + 
  geom_text(aes(x = 0.975, y = -50), label = "control", angle = 270, color = "white")
```


---

## ATE in R

* Average Treatment Effect (ATE)
  * $w_{ATE} = \frac{Z_i}{p_i} + \frac{1-Z_i}{1 - p_i}$
  
```{r}
df <- propensity_model %>% 
  augment(type.predict = "response", data = nhefs_complete) %>% 
  mutate(w_ate = (qsmk / .fitted) +  ((1 - qsmk) / (1 - .fitted))) #<<
```

---
class: inverse

## Your Turn

`r countdown::countdown(minutes = 5)`

1. Using the propensity scores you created in the previous exercise, add the ATE weights to your data frame `df`

2. Stretch: Using the same propensity scores, create ATT weights
