---
title: "Causal Modeling in R: Whole Game"
author: "Malcolm Barrett"
institute: "RStudio, PBC"
date: "2021-09-01 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
        scroll: false 
      
---

class: inverse

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, tibble.max_extra_cols = 6, tibble.width = 60)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center", dpi = 320, fig.height = 4)
library(tidyverse)
library(smd)
library(broom)
library(rsample)
library(ggdag)
library(causaldata)
```

# Broad strokes
1. Specify causal question
2. Draw assumptions (causal diagram)
3. Model assumptions (propensity score)
4. Analyze propensities (diagnostics)
5. Estimate causal effects (IPW) 

---

class: middle, center, inverse

# **Do people who quit smoking gain weight?**

---

```{r}
library(causaldata)
nhefs_complete_uc <- nhefs_complete %>% 
  filter(censored == 0)
nhefs_complete_uc
```

---

## Did those who quit smoking gain weight?

```{r, echo = FALSE}
colors <- c("#E69F00", "#56B4E9")

nhefs_complete_uc %>%
  ggplot(aes(wt82_71, fill = factor(qsmk))) + 
  geom_vline(xintercept = 0, color = "grey60", size = 1) +
  geom_density(color = "white", alpha = .75, size = .5) +
  scale_fill_manual(values = colors) + 
  theme_minimal() +
  theme(legend.position = "bottom") + 
  labs(
    x = "change in weight (kg)",
    fill = "quit smoking (1 = yes)"
  )
```

---

## Did those who quit smoking gain weight?

```{r, highlight.output = 4:5}
# ~2.5 KGs gained for quit vs. not quit
nhefs_complete_uc %>% 
  group_by(qsmk) %>% 
  summarize(
    mean_weight_change = mean(wt82_71), 
    sd = sd(wt82_71),
    .groups = "drop"
  )
```

---

class: inverse, center, middle

# **draw your assumptions**

---

```{r, echo = FALSE, fig.height=5.5}
library(ggdag)
set.seed(1234)
# set up DAG
smk_wt_dag <- dagify(
  # specify causes of quitting smoking and weight gain:
  qsmk ~ sex + race + age + education + 
    smokeintensity + smokeyrs + exercise + active + wt71,
  wt82_71 ~ qsmk + sex + race + age + education + 
    smokeintensity + smokeyrs + exercise + active + wt71,
  # specify causal question:
  exposure = "qsmk", 
  outcome = "wt82_71",
  # set up labels:
  # here, I'll use the same variable names as the data set, but I'll label them
  # with clearer names
  labels = c(
    # causal question
    "qsmk" = "quit\nsmoking",
    "wt82_71" = "change in\nweight",
    
    # demographics
    "age" = "age",
    "sex" = "sex",
    "race" = "race",
    "education" = "education",
    
    # health
    "wt71" = "baseline\nweight",
    "active" = "daily\nactivity\nlevel",
    "exercise" = "exercise",
    
    # smoking history
    "smokeintensity" = "smoking\nintensity",
    "smokeyrs" = "yrs of\nsmoking"
  )
) %>% 
  tidy_dagitty()

smk_wt_dag %>% 
  filter(name %in% c("qsmk", "wt82_71")) %>% 
  ggdag(text = FALSE, use_labels = "label") +
  ylim(6, 9.5) +
  xlim(1.5, 5)
```

---

```{r, echo = FALSE, fig.height=5.5}
smk_wt_dag %>% 
  ggdag(text = FALSE, use_labels = "label") +
  ylim(6, 9.5) +
  xlim(1.5, 5)
```

---

class:  center, middle

# What do I need to control for?

---

```{r, echo = FALSE, fig.height = 5.5}
smk_wt_dag %>% 
  ggdag_adjustment_set(text = FALSE, use_labels = "label", node_size = 8)
```

---

## Multivariable regression: what's the association?

```{r, eval = FALSE}
lm( #<<
  wt82_71~ qsmk + sex +  #<<
    race + age + I(age^2) + education + #<<
    smokeintensity + I(smokeintensity^2) + #<<
    smokeyrs + I(smokeyrs^2) + exercise + active + #<<
    wt71 + I(wt71^2), #<<
  data = nhefs_complete_uc #<<
) %>% #<<
  tidy(conf.int = TRUE) %>% 
  filter(term == "qsmk")
```

---

## Multivariable regression: what's the association?

```{r, highlight.output = 4}
lm( 
  wt82_71~ qsmk + sex +  
    race + age + I(age^2) + education + 
    smokeintensity + I(smokeintensity^2) + 
    smokeyrs + I(smokeyrs^2) + exercise + active + 
    wt71 + I(wt71^2), 
  data = nhefs_complete_uc 
) %>% 
  tidy(conf.int = TRUE) %>% 
  filter(term == "qsmk")
```

---

class: inverse, center, middle

# **model your assumptions**

---

class: center, middle

# counterfactual: what if <u>everyone</u> quit smoking vs. what if <u>no one</u> quit smoking

---

## Fit propensity score model

```{r}
propensity_model <- glm( #<<
  qsmk ~ sex +  #<<
    race + age + I(age^2) + education + 
    smokeintensity + I(smokeintensity^2) + 
    smokeyrs + I(smokeyrs^2) + exercise + active + 
    wt71 + I(wt71^2), 
  family = binomial(), 
  data = nhefs_complete_uc
)
```

---

## Calculate inverse probability weights

```{r}
nhefs_complete_uc <- propensity_model %>% 
  # predict whether quit smoking
  augment(type.predict = "response", data = nhefs_complete_uc) %>% #<<
  # calculate inverse probability
  mutate(wts = 1 / ifelse(qsmk == 0, 1 - .fitted, .fitted)) #<<
```

---

class: inverse, center, middle

# **diagnose your model assumptions**

---

## What's the distribution of weights?

```{r, echo = FALSE}
ggplot(nhefs_complete_uc, aes(wts)) +
  geom_density(col = "#E69F00", fill = "#E69F0095", size = .8) + 
  #  use a log scale for the x axis
  scale_x_log10() + 
  theme_minimal(base_size = 20) + 
  xlab("Weights")
```


---

```{r, echo=FALSE, fig.height=5.5}
vars <- c("sex", "race", "age", "education", "smokeintensity", "smokeyrs", 
          "exercise", "active", "wt71")

plot_df <- nhefs_complete_uc %>% 
  summarise(
    across(
      all_of(vars), 
      list(
        unweighted = ~smd(.x, qsmk)$estimate, 
        weighted = ~smd(.x, qsmk, wts)$estimate
      )
    )
  ) %>% 
  pivot_longer(
    everything(), 
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_sep = "_"
  )
```


```{r, echo=FALSE, fig.height=5.5}
ggplot(
  data = plot_df %>% filter(Method == "unweighted"),
  aes(x = abs(SMD), y = variable, group = Method, color = Method)
) +
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, color = "black", size = 0.1) +
  theme_minimal() +
  xlim(0, .3)
```

---

```{r, echo=FALSE, fig.height=5.5}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, group = Method, color = Method)
) +
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, color = "black", size = 0.1) +
  theme_minimal() +
  scale_color_manual(values = c("grey85", "#00BFC4")) +
  xlim(0, .3)
```

---
class: inverse, center, middle

# **estimate the causal effects**

---

## Estimate causal effect with IPW

```{r}
ipw_model <- lm( #<<
  wt82_71 ~ qsmk, #<<
  data = nhefs_complete_uc, 
  weights = wts #<<
) 

ipw_estimate <- ipw_model %>% 
  tidy(conf.int = TRUE) %>% 
  filter(term == "qsmk")
```

---

## Estimate causal effect with IPW

```{r, highlight.output = 4}
ipw_estimate
```

---

## Let's fix our confidence intervals with robust SEs!

--

```{r}
# also see robustbase, survey, gee, and others
library(estimatr)
ipw_model_robust <- lm_robust( #<<
  wt82_71 ~ qsmk, #<<
  data = nhefs_complete_uc, 
  weights = wts #<<
) 

ipw_estimate_robust <- ipw_model_robust %>% 
  tidy(conf.int = TRUE) %>% 
  filter(term == "qsmk")
```

---

## Let's fix our confidence intervals with robust SEs!

```{r, highlight.output = 4}
as_tibble(ipw_estimate_robust)
```

---

## Let's fix our confidence intervals with the bootstrap!

--

```{r}
# fit ipw model for a single bootstrap sample
fit_ipw_not_quite_rightly <- function(split, ...) { 
  # get bootstrapped data sample with `rsample::analysis()`
  .df <- analysis(split)
  
  # fit ipw model
  lm(wt82_71 ~ qsmk, data = .df, weights = wts) %>% 
    tidy()
}
```

---

```{r}
fit_ipw <- function(split, ...) {
  .df <- analysis(split)
  
  # fit propensity score model
  propensity_model <- glm(
    qsmk ~ sex + 
      race + age + I(age^2) + education + 
      smokeintensity + I(smokeintensity^2) + 
      smokeyrs + I(smokeyrs^2) + exercise + active + 
      wt71 + I(wt71^2), 
    family = binomial(), 
    data = .df
  )
  
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = 1 / ifelse(qsmk == 0, 1 - .fitted, .fitted))
  
  # fit correctly bootsrapped ipw model
  lm(wt82_71 ~ qsmk, data = .df, weights = wts) %>% 
    tidy()
}
```

---

## Using {rsample} to bootstrap our causal effect

--

```{r boot_cache, cache = TRUE}
# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(nhefs_complete, 1000, apparent = TRUE) %>% #<<
  mutate(results = map(splits, fit_ipw)) #<<
```

---

## Using {rsample} to bootstrap our causal effect

```{r, eval = FALSE}
# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, results) %>%  #<<
  filter(term == "qsmk")

boot_estimate
```

---

## Using {rsample} to bootstrap our causal effect

```{r, highlight.output = 4}
# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, results) %>% 
  filter(term == "qsmk")

boot_estimate
```

---

class: middle

```{r, echo = FALSE}
bind_rows(
  ipw_estimate %>% 
    select(estimate, conf.low, conf.high) %>% 
    mutate(type = "ols"),
  ipw_estimate_robust %>% 
    select(estimate, conf.low, conf.high) %>% 
    mutate(type = "robust"),
  boot_estimate %>% 
    select(estimate = .estimate, conf.low = .lower, conf.high = .upper) %>% 
    mutate(type = "bootstrap")
) %>% 
  #  calculate CI width to sort by it
  mutate(width = conf.high - conf.low) %>% 
  arrange(width) %>% 
  #  fix the order of the model types for the plot  
  mutate(type = fct_inorder(type)) %>% 
  ggplot(aes(x = type, y = estimate, ymin = conf.low, ymax = conf.high)) + 
    geom_pointrange(color = "#0172B1", size = 1, fatten = 3) +
    coord_flip() +
    theme_minimal(base_size = 20) +
    theme(axis.title.y = element_blank())
```

---

class: center, inverse, middle

# *Our causal effect estimate: **3.5 kg (95% CI 2.4 kg, 4.4 kg)***

---

class: center, inverse, middle

# **Review the R Markdown file... later!**

---
class: inverse, center

# Resources
## [Causal Inference](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/): Comprehensive text on causal inference. Free online.
## [Causal Inference Notebook](http://causalinferencebookr.netlify.com): R code to go along with Causal Inference
## [Bootstrap confidence intervals with {rsample}](https://rsample.tidymodels.org/articles/Applications/Intervals.html)
