---
title: "Propensity Score Diagnostics"
author: "Lucy D'Agostino McGowan"
institute: "Wake Forest University"
date: "2021-09-01 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      navigation:
        scroll: false 
---

```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center", dpi = 320, fig.height = 4)
```

```{r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

class: inverse

## Checking balance

* Love plots (Standardized Mean Difference)
* ECDF plots

---
class: inverse

## Standardized Mean Difference (SMD)

$$\LARGE d = \frac{\bar{x}_{treatment}-\bar{x}_{control}}{\sqrt{\frac{s^2_{treatment}+s^2_{control}}{2}}}$$

---

## SMD in R

```{r, message = FALSE, warning = FALSE}
library(smd)
library(tidyverse)

df %>% 
  # w is optional
  summarise(smd = smd(z, x, w = wts))
```

---

## SMD in R

<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Calculate standardized mean differences </h3> 


```{r,  message = FALSE, warning = FALSE}
smds <- df %>% 
  summarise( 
    across( 
      z1, z2, z3, 
      list(
        unweighted = ~smd(.x, x)$estimate, 
        weighted = ~smd(.x, x, wts)$estimate 
      )
    )
  )
```

---

## SMD in R

<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Calculate standardized mean differences </h3> 


```{r,  message = FALSE, warning = FALSE}
smds <- df %>% 
  summarise( #<<
    across( #<<
      z1, z2, z3, #<<
      list(
        unweighted = ~smd(.x, x)$estimate, 
        weighted = ~smd(.x, x, wts)$estimate 
      )
    )
  )
```

---

## SMD in R

<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Calculate standardized mean differences </h3> 


```{r,  message = FALSE, warning = FALSE}
smds <- df %>% 
  summarise(
    across(
      z1, z2, z3,
      list(
        unweighted = ~smd(.x, x)$estimate, #<<
        weighted = ~smd(.x, x, wts)$estimate #<<
      )
    )
  )
```

---

## SMD in R

<span class = "num">3</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Pivot the SMDs into tidy format </h3> 

```{r}
plot_df <- smds %>% 
  pivot_longer( 
    everything(),
    values_to = "SMD", 
    names_to = c("variable", "Method"), 
    names_sep = "_"
  )
```

---

## SMD in R

<span class = "num">3</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Pivot the SMDs into tidy format </h3> 

```{r}
plot_df <- smds %>% 
  pivot_longer( #<<
    everything(), #<<
    values_to = "SMD", 
    names_to = c("variable", "Method"), #<<
    names_sep = "_"
  )
```

---

## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, color = "black", size = 0.1)
```

---

## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(
  data = plot_df, #<<
  aes(x = abs(SMD), y = variable, group = Method, color = Method) #<<
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, color = "black", size = 0.1)
```

---


## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, group = Method, color = Method)
) +  
  geom_line(orientation = "y") + #<<
  geom_point() + 
  geom_vline(xintercept = 0.1, color = "black", size = 0.1)
```

---


## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, group = Method, color = Method)
) +
  geom_line(orientation = "y") +
  geom_point() + #<<
  geom_vline(xintercept = 0.1, color = "black", size = 0.1)
```


---


## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = variable, group = Method, color = Method)
) +
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, color = "black", size = 0.1) #<<
```

---

## Love plot

```{r, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}
library(smd)
library(tidyverse)
library(broom)
library(causaldata)

propensity_model <- glm(
  qsmk ~ sex + 
    race + age + I(age^2) + education + 
    smokeintensity + I(smokeintensity^2) + 
    smokeyrs + I(smokeyrs^2) + exercise + active + 
    wt71 + I(wt71^2), 
  family = binomial(), 
  data = nhefs_complete
)

df <- propensity_model %>% 
  augment(type.predict = "response", data = nhefs_complete) %>% 
  mutate(w_ate = 1 / ifelse(qsmk == 0, 1 - .fitted, .fitted))

vars <- c("sex", "race", "age", "education", "smokeintensity", "smokeyrs", 
          "exercise", "active", "wt71")

plot_df <- df %>% 
  summarise(
    across(
      all_of(vars), 
      list(
        unweighted = ~smd(.x, qsmk)$estimate, 
        weighted = ~smd(.x, qsmk, w_ate)$estimate
      )
    )
  ) %>% 
  pivot_longer(
    everything(), 
    values_to = "SMD", 
    names_to = c("variable", "Method"),
    names_sep = "_"
  )

ggplot(
  data = plot_df,
  mapping = aes(x = abs(SMD), y = variable, group = Method, color = Method)
) +
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, color = "black", size = 0.1)
```

---

## Your turn 1

`r countdown::countdown(minutes = 10)`

1. Create a Love Plot for the propensity score weighting you created in the previous exercise

---

## ECDF

For continuous variables, it can be helpful to look at the _whole_ distribution pre and post-weighting rather than a single summary measure

```{r, echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) +
  stat_ecdf() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```


---

## Unweighted ECDF

```{r}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) +
  stat_ecdf() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Unweighted ECDF

```{r}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) + #<<
  stat_ecdf() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Unweighted ECDF

```{r}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) +
  stat_ecdf() + #<<
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Unweighted ECDF

```{r, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) +
  stat_ecdf() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```


---

## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>% #<<
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>% #<<
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate)) #<<

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>% #<<
  filter(qsmk == 0) %>% #<<
  arrange(wt71) %>% #<<
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate)) #<<

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) + #<<
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") + #<<
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") +  #<<
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```
---

## Weighted  ECDF 

```{r, echo = FALSE, eval = TRUE}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Your turn 2

`r countdown::countdown(minutes = 10)`

1. Create an unweighted ECDF examining the `smokeyrs` confounder for those that quit smoking and those that did not
3. Create a weighted ECDF examining the `smokeyrs` confounder


---

class: middle, center, inverse

## *Bonus!* Weighted Tables in R


---

## Weighted Tables in R

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Create a "design object" to incorporate the weights </h3> 

```{r, message = FALSE, warning = FALSE}
library(survey)

svy_des <- svydesign(
  ids = ~ 1,
  data = df,
  weights = ~ wts,
  strata = ~ x
)
```

---

## Weighted Tables in R

<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Pass to `gtsummary::tbl_svysummary()` </h3> 

```{r,  message = FALSE, warning = FALSE}
library(gtsummary)
tbl_svysummary(svy_des, by = x) %>% 
  add_difference(everything() ~ "smd")
```

---

```{r, message = FALSE, warning = FALSE, eval = TRUE, echo = FALSE}
library(survey)
library(gtsummary)

svy_des <- svydesign(
  ids = ~ 1,
  data = df %>% select(wt71, race, age, sex, smokeintensity, smokeyrs, exercise, active, w_ate, qsmk),
  weights = ~ w_ate
)
theme_gtsummary_compact()

tbl_svysummary(svy_des, by = qsmk) %>% 
  add_difference(everything() ~ "smd")
```
