---
title: "Propensity Score Diagnostics"
author: "Lucy D'Agostino McGowan"
institute: "Wake Forest University"
date: "2021-09-01 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      navigation:
        scroll: false 
---

```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center", dpi = 320, fig.height = 4)
```

```{r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

class: inverse

## Checking balance

* Love plots (Standardized Mean Difference)
* ECDF plots

---
class: inverse

## Standardized Mean Difference (SMD)

$$\LARGE d = \frac{\bar{x}_{treatment}-\bar{x}_{control}}{\sqrt{\frac{s^2_{treatment}+s^2_{control}}{2}}}$$

---

## SMD in R

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Create a "design object" to incorporate the weights </h3> 

```{r, message = FALSE, warning = FALSE}
library(survey)

svy_des <- svydesign(
  ids = ~ 1,
  data = df,
  weights = ~ wts
)
```

---

## SMD in R

<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Calculate the unweighted standardized mean differences </h3> 


```{r,  message = FALSE, warning = FALSE}
library(tableone)
library(tidyverse)

smd_table_unweighted <- CreateTableOne(
  vars = c("confounder_1", "confounder_1", ...),
  strata = "exposure",
  data = df,
  test = FALSE)
```

---

## SMD in R

<span class = "num">3</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Calculate the weighted standardized mean differences </h3> 

```{r}
smd_table <- svyCreateTableOne(
  vars = c("confounder_1", "confounder_1", ...),
  strata = "exposure",
  data = svy_des, 
  test = FALSE)
```

---

## SMD in R

<span class = "num">3</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Calculate the weighted standardized mean differences </h3> 

```{r}
smd_table <- svyCreateTableOne( #<<
  vars = c("confounder_1", "confounder_1", ...),
  strata = "exposure",
  data = svy_des, #<<
  test = FALSE)
```

---

## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Stick these together in a data frame </h3> 


```{r}
plot_df <- data.frame(
  var = rownames(ExtractSmd(smd_table)),                        
  Unadjusted = as.numeric(ExtractSmd(smd_table_unweighted)),                      
  Weighted = as.numeric(ExtractSmd(smd_table))) %>%
  pivot_longer(-var, names_to = "Method", values_to = "SMD")
```

---

## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Stick these together in a data frame </h3> 


```{r}
plot_df <- data.frame(
  var = rownames(ExtractSmd(smd_table)), #<<                        
  Unadjusted = as.numeric(ExtractSmd(smd_table_unweighted)),                      
  Weighted = as.numeric(ExtractSmd(smd_table))) %>%
  pivot_longer(-var, names_to = "Method", values_to = "SMD")

rownames(EXtractSMD(smd_table))
#> [1] "confounder_1"  "confounder_2"
```

---

## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Stick these together in a data frame </h3> 


```{r}
plot_df <- data.frame(
  var = rownames(ExtractSmd(smd_table)),                        
  Unadjusted = as.numeric(ExtractSmd(smd_table_unweighted)),  #<<                    
  Weighted = as.numeric(ExtractSmd(smd_table))) %>%
  pivot_longer(-var, names_to = "Method", values_to = "SMD")

as.numeric(ExtractSmd(smd_table_unweighted))
#> [1] 0.160 0.177
```

---

## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Stick these together in a data frame </h3> 


```{r}
plot_df <- data.frame(
  var = rownames(ExtractSmd(smd_table)),                        
  Unadjusted = as.numeric(ExtractSmd(smd_table_unweighted)),                    
  Weighted = as.numeric(ExtractSmd(smd_table))) %>% #<<
  pivot_longer(-var, names_to = "Method", values_to = "SMD")

as.numeric(ExtractSmd(smd_table))
#> [1] 0.002 0.007
```

---

## SMD in R

<span class = "num">4</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Stick these together in a data frame </h3> 


```{r}
plot_df <- data.frame(
  var = rownames(ExtractSmd(smd_table)),                        
  Unadjusted = as.numeric(ExtractSmd(smd_table_unweighted)),                    
  Weighted = as.numeric(ExtractSmd(smd_table))) %>%
  pivot_longer(-var, names_to = "Method", values_to = "SMD") #<<

```

---

## SMD in R

<span class = "num">5</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(data = plot_df, 
       mapping = aes(x = var, y = SMD, group = Method, color = Method)) +
  geom_line() +
  geom_point() + 
  geom_hline(yintercept = 0.1, color = "black", size = 0.1) +  
  coord_flip()
```

---

## SMD in R

<span class = "num">5</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(data = plot_df, #<<
       mapping = aes(x = var, y = SMD, group = Method, color = Method)) + #<<
  geom_line() +
  geom_point() + 
  geom_hline(yintercept = 0.1, color = "black", size = 0.1) +  
  coord_flip()
```

---


## SMD in R

<span class = "num">5</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(data = plot_df,
       mapping = aes(x = var, y = SMD, group = Method, color = Method)) + 
  geom_line() + #<<
  geom_point() + 
  geom_hline(yintercept = 0.1, color = "black", size = 0.1) +  
  coord_flip()
```

---


## SMD in R

<span class = "num">5</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(data = plot_df,
       mapping = aes(x = var, y = SMD, group = Method, color = Method)) + 
  geom_line() +
  geom_point() + #<<
  geom_hline(yintercept = 0.1, color = "black", size = 0.1) +  
  coord_flip()
```


---


## SMD in R

<span class = "num">5</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 


```{r}
ggplot(data = plot_df,
       mapping = aes(x = var, y = SMD, group = Method, color = Method)) + 
  geom_line() +
  geom_point() + 
  geom_hline(yintercept = 0.1, color = "black", size = 0.1) +  #<<
  coord_flip()
```


---


## SMD in R


<span class = "num">5</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp;  Plot them! (in a Love plot!) </h3> 

```{r}
ggplot(data = plot_df, mapping = aes(x = var, y = SMD, group = Method, color = Method)) +
  geom_line() +
  geom_point() + 
  geom_hline(yintercept = 0.1, color = "black", size = 0.1) +  
  coord_flip() #<<
```

---

## Love plot

```{r, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}
library(survey)
library(tableone)
library(tidyverse)
library(broom)
library(causaldata)

propensity_model <- glm(
  qsmk ~ sex + 
    race + age + I(age^2) + education + 
    smokeintensity + I(smokeintensity^2) + 
    smokeyrs + I(smokeyrs^2) + exercise + active + 
    wt71 + I(wt71^2), 
  family = binomial(), 
  data = nhefs_complete
)

df <- propensity_model %>% 
  augment(type.predict = "response", data = nhefs_complete) %>% 
  mutate(w_ate = 1 / ifelse(qsmk == 0, 1 - .fitted, .fitted))

svy_des <- svydesign(
  ids = ~ 1,
  data = df,
  weights = ~ w_ate)

smd_table_unweighted <- CreateTableOne(
  vars = c("sex", "race", "age", "education", "smokeintensity", "smokeyrs", 
           "exercise", "active", "wt71"),
  strata = "qsmk",
  data = df,
  test = FALSE)

smd_table <- svyCreateTableOne(
  vars = c("sex", "race", "age", "education", "smokeintensity", "smokeyrs", 
           "exercise", "active", "wt71"),
  strata = "qsmk",
  data = svy_des,
  test = FALSE)


plot_df <- data.frame(
  var = rownames(ExtractSmd(smd_table)),                        
  Unadjusted = as.numeric(ExtractSmd(smd_table_unweighted)),                      
  Weighted = as.numeric(ExtractSmd(smd_table))) %>%
  pivot_longer(-var, names_to = "Method", values_to = "SMD")

ggplot(
  data = plot_df,
  mapping = aes(x = var, y = SMD, group = Method, color = Method)
) +
  geom_line() +
  geom_point() + 
  geom_hline(yintercept = 0.1, color = "black", size = 0.1) +  
  coord_flip()
```

---

## Your turn 1

`r countdown::countdown(minutes = 10)`

1. Create a Love Plot for the propensity score weighting you created in the previous exercise

---

## ECDF

For continuous variables, it can be helpful to look at the _whole_ distribution pre and post-weighting rather than a single summary measure

```{r, echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) +
  stat_ecdf() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```


---

## Unweighted ECDF

```{r}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) +
  stat_ecdf() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Unweighted ECDF

```{r}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) + #<<
  stat_ecdf() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Unweighted ECDF

```{r}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) +
  stat_ecdf() + #<<
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Unweighted ECDF

```{r, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}
ggplot(df, aes(x = wt71, group = qsmk, color = factor(qsmk))) +
  stat_ecdf() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```


---

## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>% #<<
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>% #<<
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate)) #<<

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>% #<<
  filter(qsmk == 0) %>% #<<
  arrange(wt71) %>% #<<
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate)) #<<

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) + #<<
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") + #<<
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---


## Weighted  ECDF 

```{r}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") +  #<<
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```
---

## Weighted  ECDF 

```{r, echo = FALSE, eval = TRUE}
ecdf_1 <- df %>%
  filter(qsmk == 1) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(qsmk == 0) %>%
  arrange(wt71) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = wt71, y = cum_pct)) +
  geom_line( color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = wt71, y = cum_pct), color = "#5154B8") + 
  xlab("Weight in Kg in 1971") + 
  ylab("Proportion <= x") 
```

---

## Your turn 2

`r countdown::countdown(minutes = 10)`

1. Create an unweighted ECDF examining the `smokeyrs` confounder for those that quit smoking and those that did not
3. Create a weighted ECDF examining the `smokeyrs` confounder
