---
title: "Fitting the outcome model"
author: "Lucy D'Agostino McGowan"
institute: "Wake Forest University"
date: "2021-09-01 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      navigation:
        scroll: false 
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Outcome Model

```{r}
library(broom)

lm(outcome ~ exposure, data = df, weights = wts) %>% 
  tidy()
```

--
`r emo::ji("check")` This will get us the point estimate  
--

`r emo::ji("x")` This will get NOT us the correct confidence intervals  
--

`r emo::ji("package")` {rsample}
---

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Create a function to run your analysis once on a sample of your data</h3> 

.small[
```{r}
fit_ipw <- function(split, ...) {
  .df <- analysis(split)
  
  # fit propensity score model
  propensity_model <- glm(
    exposure ~ confounder_1 + confounder_2 + ...
    family = binomial(), 
    data = .df
  )
  
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = 1 / ifelse(exposure == 0, 1 - .fitted, .fitted))
  
  # fit correctly bootsrapped ipw model
  lm(outcome ~ exposure, data = .df, weights = wts) %>% 
    tidy()
}
```
]

---

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Create a function to run your analysis once on a sample of your data</h3> 

.small[
```{r}
fit_ipw <- function(split, ...) { #<<
  .df <- analysis(split) #<<
  
  # fit propensity score model
  propensity_model <- glm(
    exposure ~ confounder_1 + confounder_2 + ...
    family = binomial(), 
    data = .df
  )
  
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = 1 / ifelse(exposure == 0, 1 - .fitted, .fitted))
  
  # fit correctly bootsrapped ipw model
  lm(outcome ~ exposure, data = .df, weights = wts) %>% 
    tidy()
}
```
]

---

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Create a function to run your analysis once on a sample of your data</h3> 

.small[
```{r}
fit_ipw <- function(split, ...) {
  .df <- analysis(split)
  
  # fit propensity score model #<<
  propensity_model <- glm( #<<
    exposure ~ confounder_1 + confounder_2 + ... #<<
    family = binomial(),  #<<
    data = .df #<<
  ) #<<
  
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = 1 / ifelse(exposure == 0, 1 - .fitted, .fitted))
  
  # fit correctly bootsrapped ipw model
  lm(outcome ~ exposure, data = .df, weights = wts) %>% 
    tidy()
}
```
]

---

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Create a function to run your analysis once on a sample of your data</h3> 

.small[
```{r}
fit_ipw <- function(split, ...) { 
  .df <- analysis(split) 
  
  # fit propensity score model
  propensity_model <- glm(
    exposure ~ confounder_1 + confounder_2 + ...
    family = binomial(), 
    data = .df
  )
  
  # calculate inverse probability weights #<<
  .df <- propensity_model %>%  #<<
    augment(type.predict = "response", data = .df) %>%  #<<
    mutate(wts = 1 / ifelse(exposure == 0, 1 - .fitted, .fitted)) #<<
  
  # fit correctly bootsrapped ipw model
  lm(outcome ~ exposure, data = .df, weights = wts) %>% 
    tidy() 
}
```
]

---

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Create a function to run your analysis once on a sample of your data</h3> 

.small[
```{r}
fit_ipw <- function(split, ...) { 
  .df <- analysis(split) 
  
  # fit propensity score model
  propensity_model <- glm(
    exposure ~ confounder_1 + confounder_2 + ...
    family = binomial(), 
    data = .df
  )
  
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    augment(type.predict = "response", data = .df) %>% 
    mutate(wts = 1 / ifelse(exposure == 0, 1 - .fitted, .fitted))
  
  # fit correctly bootsrapped ipw model #<<
  lm(outcome ~ exposure, data = .df, weights = wts) %>% #<<
    tidy() #<<
}
```
]

---

<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Use {rsample} to bootstrap our causal effect</h3> 


```{r}
library(rsample)

# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(df, 1000, apparent = TRUE) %>% 
  mutate(results = map(splits, fit_ipw)) 
```

---


<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Use {rsample} to bootstrap our causal effect</h3> 


```{r}
library(rsample)

# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(df, 1000, apparent = TRUE) %>%  #<<
  mutate(results = map(splits, fit_ipw)) 
```

---

<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Use {rsample} to bootstrap our causal effect</h3> 


```{r}
library(rsample)

# fit ipw model to bootstrapped samples
ipw_results <- bootstraps(df, 1000, apparent = TRUE) %>% 
  mutate(results = map(splits, fit_ipw)) #<<
```

---

<span class = "num">3</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; Pull out the causal effect</h3> 


```{r, eval = FALSE}
# get t-statistic-based CIs
boot_estimate <- int_t(ipw_results, results) %>%  #<<
  filter(term == "exposure")
```

---

## Your Turn

`r countdown::countdown(minutes = 12)`

1. Create a function called `ipw_fit` that fits the propensity score model and the weighted outcome model for the effect between `extra_magic_morning` and `avg_spostmin`

2. Using the `bootstraps()` and `int_t()` functions to estimate the final effect.


