---
title: "Tipping Point Sensitivity Analyses"
author: "Lucy D'Agostino McGowan"
institute: "Wake Forest University"
date: "2021-09-01 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: ["default", "theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      navigation:
        scroll: false 
---

```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.align = "center", dpi = 320, fig.height = 4)
```

```{r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

class: inverse

## Recall: Propensity scores

Rosenbaum and Rubin showed in observational studies, conditioning on **propensity scores** can lead to unbiased estimates of the exposure effect

1. There are no unmeasured confounders
2. Every subject has a nonzero probability of receiving either exposure

---
## Quantifying Unmeasured Confounding

### What you'll need:

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The exposure-outcome effect</h3>
<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured counfounder-exposure effect</h3> 
<span class = "num">3</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured confounder-outcome effect</h3>

---

class: center, middle

## What will tip our confidence bound to cross zero?

---

## Quantifying Unmeasured Confounding

### What you'll need:

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The exposure-outcome effect</h3>
<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured counfounder-exposure effect</h3> 
<span class = "num" style="color:grey">3</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured confounder-outcome effect</h3>

---

## Quantifying Unmeasured Confounding

### What you'll need:

<span class = "num">1</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The exposure-outcome effect</h3>
<span class = "num" style="color:grey">2</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured counfounder-exposure effect</h3> 
<span class = "num">3</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured confounder-outcome effect</h3>

---

## Tipping point

$$\Huge \beta_{UO}(LB_{obs}, \delta)$$

---

## Tipping point

$$\Huge \beta_{UO}(LB_{obs}, \delta)$$

* $\beta_{UO}$: the **unmeasured confounder-outcome effect**
---

## Quantifying Unmeasured Confounding

### What you'll need:

<span class = "num" style="color:grey">1</span> <h3 style="color:grey">&nbsp; &nbsp; &nbsp;  &nbsp; The exposure-outcome effect</h3>
<span class = "num" style="color:grey">2</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured counfounder-exposure effect</h3> 
<span class = "num">3</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured confounder-outcome effect</h3>

---

## Tipping point

$$\Huge \beta_{UO}(LB_{obs}, \delta)$$

* $LB_{obs}$: **limiting bound** - the bound closest to the null
---

## Quantifying Unmeasured Confounding

### What you'll need:

<span class = "num" >1</span> <h3>&nbsp; &nbsp; &nbsp;  &nbsp; The exposure-outcome effect</h3>
<span class = "num" style="color:grey">2</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured counfounder-exposure effect</h3> 
<span class = "num" style="color:grey">3</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured confounder-outcome effect</h3>

---

## Tipping point

$$\Huge \beta_{UO}(LB_{obs}, \delta)$$

* $\delta$: **standardized mean difference** of the unmeasured confounder between the exposed and unexposed groups

---

## Quantifying Unmeasured Confounding

### What you'll need:

<span class = "num"  style="color:grey">1</span> <h3 style="color:grey">&nbsp; &nbsp; &nbsp;  &nbsp; The exposure-outcome effect</h3>
<span class = "num">2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured counfounder-exposure effect</h3> 
<span class = "num" style="color:grey">3</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured confounder-outcome effect</h3>

---
## Tipping Point

$$\Huge \beta_{UO}(LB_{obs}, \delta)=\frac{LB_{obs}}{\delta}$$
---


## Tipping Point

$$\Huge \delta(LB_{obs}, \beta_{UO})=\frac{LB_{obs}}{\beta_{UO}}$$

---

<!-- ## Tipping Point -->

<!-- $$\Huge RR_{UO}(LB_{obs}, \delta)=LB_{obs}^{1/\delta}$$ -->

<!-- ## Tipping Point -->

<!-- $$\Huge \delta(LB_obs, RR_{UO})=\frac{\log(LB_{obs})}{\log(RR_{UO})}$$ -->

---

<!-- ## Tipping Point -->

<!-- $$\Huge n(LB_{obs}, RR_{UO}, \delta) = \frac{\log{LB_{obs}}}{\delta\log{RR_{UO}}}$$ -->
---

class: inverse, center, middle

# tipr

---

## Main function

# `tip_coef()`

* `effect`: observed exposure - outcome effect 
---


## Quantifying Unmeasured Confounding

### What you'll need:

<span class = "num">1</span> <h3>&nbsp; &nbsp; &nbsp;  &nbsp; The exposure-outcome effect</h3>
<span class = "num" style="color:grey">2</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured counfounder-exposure effect</h3> 
<span class = "num" style="color:grey">3</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured confounder-outcome effect</h3>
---

## Main function

# `tip_coef()`
* `smd`: scaled mean difference between the unmeasured confounder in the exposed and unexposed population


---

## Quantifying Unmeasured Confounding

### What you'll need:

<span class = "num" style="color:grey" >1</span> <h3 style="color:grey">&nbsp; &nbsp; &nbsp;  &nbsp; The exposure-outcome effect</h3>
<span class = "num" >2</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured counfounder-exposure effect</h3> 
<span class = "num" style="color:grey">3</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured confounder-outcome effect</h3>

---


## Main function

# `tip_coef()`
* `outcome_association`: association between the unmeasured confounder and outcome

---

## Quantifying Unmeasured Confounding

### What you'll need:

<span class = "num"  style="color:grey">1</span> <h3 style="color:grey">&nbsp; &nbsp; &nbsp;  &nbsp; The exposure-outcome effect</h3>
<span class = "num" style="color:grey">2</span> <h3 style="color:grey"> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured counfounder-exposure effect</h3> 
<span class = "num" >3</span> <h3> &nbsp; &nbsp; &nbsp;  &nbsp; The unmeasured confounder-outcome effect</h3>
---

## Main function

# `tip_coef()`

* `effect`
* `smd`
* `outcome_association`
---

## Main function

# `tip_coef()`

### **specify** one, it will **estimate** the other

* `d`
* <span style="color:orange">`smd`</span>
* <span style="color:orange">`outcome_association` </span>

---

## Example

*Our causal effect estimate: **3.5 lbs (95% CI 2.4 lbs, 4.4 lbs)***

--

```{r, eval = TRUE}
library(tipr)
tip_coef(effect = 2.4,
         smd = 0.3)
```

--

The observed effect (2.4, 4.4) WOULD be tipped by 1 unmeasured confounder
with the following specifications:

*estimated standardized mean difference between the unmeasured confounder
in the exposed population and unexposed population: 0.3*  

*estimated association between the unmeasured confounder and the outcome: 8*
---

## Your turn

`r countdown::countdown(minutes = 10)`

1. Use the `tip_coef()` function to conduct a sensitivity analysis for the estimate from your previous exercises.
