---
title: "Continuous exposures with propensity scores"
author: "Malcolm Barrett"
format: "kakashi-revealjs"
date: "2021-09-01 (updated: `r Sys.Date()`)"
---

```{r}
#| label: setup
#| include: false
options(
  tibble.max_extra_cols = 6, 
  tibble.width = 60
)

library(tidyverse)
library(broom)
library(causaldata)
library(touringplans)
library(propensity)

set.seed(1000)
```

## {background-color="#23373B" .huge .center}
### *Warning!* Propensity score weights are sensitive to positivity violations for continuous exposures. 

## {background-color="#23373B" .huge .center}
### **The story so far**

---

## Propensity score weighting {background-color="#23373B"}

1. Fit a propensity model predicting exposure `x`, `x + z` where z is all covariates
2. Calculate weights 
3. Fit an outcome model estimating the effect of `x` on `y` weighted by the propensity score

## Continous exposures {background-color="#23373B"}

1. Use a model like `lm(x ~ z)` for the propensity score model.
2. Use `wt_ate()` with `.fitted` and `.sigma`; transforms using `dnorm()` to get on probability-like scale.
3. Apply the weights to the outcome model as normal!

## Alternative: quantile binning {background-color="#23373B" .small} 

1. Bin the continuous exposure into quantiles and use categorical regression like a multinomial model to calculate probabilities.
2. Calculate the weights where the propensity score is the probability you fall into the quantile you actually fell into. Same as the binary ATE!
3. Same workflow for the outcome model

## 1. Fit a model for `exposure ~ confounders`

```{r}
#| eval: false
model <- lm(
  exposure ~ confounder_1 + confounder_2,
  data = df
)
```

## 2. Calculate the weights with `wt_ate()`

```{r}
#| eval: false
#| code-line-numbers: "|3-8"
model |>
  augment(data = df) |>
  mutate(wts = wt_ate( 
    exposure, 
    .fitted, 
    # .sigma is from augment()
    .sigma = .sigma
  )) 
```

## Does change in smoking intensity (`smkintensity82_71`) affect weight gain among lighter smokers?

```{r}
nhefs_light_smokers <- nhefs_complete |>
  filter(smokeintensity <= 25)
```

## 1. Fit a model for `exposure ~ confounders`

```{r}
#| code-line-numbers: "|1-2|3-6"
nhefs_model <- lm(
  smkintensity82_71 ~ sex + race + age + I(age^2) + 
    education + smokeintensity + I(smokeintensity^2) + 
    smokeyrs + I(smokeyrs^2) + exercise + active + 
    wt71 + I(wt71^2), 
  data = nhefs_light_smokers
)
```

## 2. Calculate the weights with `wt_ate()`

```{r}
#| code-line-numbers: "|3-7"
nhefs_wts <- nhefs_model |> 
  augment(data = nhefs_light_smokers) |> 
  mutate(wts = wt_ate(
    smkintensity82_71, 
    .fitted,
    .sigma = .sigma
  )) 
```


## 2. Calculate the weights with `wt_ate()`

```{r}
nhefs_wts
```

## Do *posted* wait times at 8 am affect *actual* wait times at 9 am?

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig.width: 6.5
library(tidyverse)
library(ggdag)
library(ggokabeito)

geom_dag_label_repel <- function(..., seed = 10) {
  ggdag::geom_dag_label_repel(
    aes(x, y, label = label),
    box.padding = 3.5, 
    inherit.aes = FALSE,
    max.overlaps = Inf, 
    family = "sans",
    seed = seed,
    label.size = NA, 
    label.padding = 0.1,
    size = 14 / 3,
    ...
  ) 
}

coord_dag <- list(
  x = c(Season = -1, close = -1, weather = -2, extra = 0, x = 1, y = 2),
  y = c(Season = -1, close = 1, weather = 0, extra = 0, x = 0, y = 0)
)

labels <- c(
  extra = "Extra Magic Morning",
  x = "Average posted wait ",
  y = "Average acutal wait",
  Season = "Ticket Season",
  weather = "Historic high temperature",
  close = "Time park closed"
)

dagify(
    y ~ x + close + Season + weather + extra,
    x ~ weather + close + Season + extra,
    extra ~ weather + close + Season,
    coords = coord_dag,
    labels = labels,
    exposure = "x",
    outcome = "y"
) |>
    tidy_dagitty() |>
    node_status() |>
    ggplot(
        aes(x, y, xend = xend, yend = yend, color = status)
    ) +
    geom_dag_edges_arc(curvature = c(rep(0, 7), .2, 0, .2, .2, 0), edge_colour = "grey70") +
    geom_dag_point() +
    geom_dag_label_repel(seed = 1602) + 
    scale_color_okabe_ito(na.value = "grey90") +
    theme_dag() +
    theme(
        legend.position = "none",
        axis.text.x = element_text()
    ) +
    coord_cartesian(clip = "off") +
    scale_x_continuous(
        limits = c(-2.25, 2.25),
        breaks = c(-2, -1, 0, 1, 2),
        labels = c(
            "\n(one year ago)",
            "\n(6 months ago)",
            "\n(3 months ago)",
            "8am-9am\n(Today)",
            "9am-10am\n(Today)"
        )
    )
```

## *Your Turn 1*

### Fit a model using `lm()` with `wait_minutes_posted_avg` as the outcome and the confounders identified in the DAG.
### Use `augment()` to add model predictions to the data frame
### In `wt_ate()`, calculate the weights using `wait_minutes_posted_avg`, `.fitted`, and `.sigma`

`r countdown::countdown(minutes = 5)`

## *Your Turn 1*

```{r}
#| include: false
eight <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 8) |>
  select(-wait_minutes_posted_avg)

nine <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9) |>
  select(park_date, wait_minutes_posted_avg)

wait_times <- eight |>
  left_join(nine, by = "park_date") |>
  drop_na(wait_minutes_posted_avg)
```

```{r}
post_time_model <- lm(
  wait_minutes_posted_avg ~
    park_close + park_extra_magic_morning + 
    park_temperature_high + park_ticket_season, 
  data = wait_times
)
```

## *Your Turn 1*

```{r}
wait_times_wts <- post_time_model |>
  augment(data = wait_times) |>
  mutate(wts = wt_ate(
    wait_minutes_posted_avg, .fitted, .sigma = .sigma
  ))
```

## *Stabilizing extreme weights*

```{r}
#| echo: false
nhefs_wts |>
  ggplot(aes(wts)) +
  geom_density(col = "#E69F00", fill = "#E69F0095") + 
  scale_x_log10() + 
  theme_minimal(base_size = 20) + 
  xlab("Weights")
```

## Stabilizing extreme weights {background-color="#23373B"}

1. Fit an intercept-only model (e.g. `lm(x ~ 1)`) or use mean and SD of `x`
2. Calculate weights from this model.
3. Divide these weights by the propensity score weights. `wt_ate(.., stabilize = TRUE)` does this all!

## Calculate stabilized weights 

```{r}
#| code-line-numbers: "|7"
nhefs_swts <- nhefs_model |>
  augment(data = nhefs_light_smokers) |>
  mutate(swts = wt_ate(
    smkintensity82_71, 
    .fitted, 
    .sigma = .sigma,
    stabilize = TRUE
  ))
```

## Stabilizing extreme weights

```{r}
#| echo: false
ggplot(nhefs_swts, aes(swts)) +
  geom_density(col = "#E69F00", fill = "#E69F0095") + 
  scale_x_log10() + 
  theme_minimal(base_size = 20) + 
  xlab("Stabilized Weights")
```

## *Your Turn 2*

### Re-fit the above using stabilized weights

`r countdown::countdown(minutes = 3)`

## *Your Turn 2*

```{r}
wait_times_swts <- post_time_model |>
  augment(data = wait_times) |>
  mutate(swts = wt_ate(
    wait_minutes_posted_avg, 
    .fitted,
    .sigma = .sigma,
    stabilize = TRUE
  ))
```

## Fitting the outcome model {background-color="#23373B"}

1. Use the stabilized weights in the outcome model. Nothing new here!

---

```{r}
#| code-line-numbers: "|3,8"
lm(
  wt82_71 ~ smkintensity82_71, 
  weights = swts, 
  data = nhefs_swts
) |>
  tidy() |>
  filter(term == "smkintensity82_71") |>
  mutate(estimate = estimate * -10) 
```

## *Your Turn 3*

### Estimate the relationship between posted wait times and actual wait times using the stabilized weights we just created. 

`r countdown::countdown(minutes = 3)`

## *Your Turn 3*

```{r}
lm(
  wait_minutes_actual_avg ~ wait_minutes_posted_avg, 
  weights = swts, 
  data = wait_times_swts
) |>
  tidy() |>
  filter(term == "wait_minutes_posted_avg") |>
  mutate(estimate = estimate * 10)
```


## Diagnosing issues {background-color="#23373B"}

1. Extreme weights even after stabilization
2. Bootstrap: non-normal distribution
3. Bootstrap: estimate different from original model

## More info {background-color="#23373B"}

### https://github.com/LucyMcGowan/writing-positivity-continous-ps
