---
title: "Interaction and Heterogeneous Treatment Effects"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

## {background-color="#23373B"}

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)
library(tidyverse)
library(broom)
library(propensity)
library(marginaleffects)
library(causalgenerics)
library(nnet)
library(ggdag)

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)
```

- The ATE, ATT, and ATC are all expectations of the same unit-level contrast,
  `y(1) - y(0)`, taken over different covariate distributions
- That distinction becomes meaningful only when the individual treatment effect
  varies across units

## Two claims from earlier modules

- When standard methods succeed: the effect was homogeneous, so marginal and
  conditional effects agree
- In g-computation: without an exposure-covariate interaction, the ATE, ATT, and
  ATC coincide

. . .

This module is where both claims stop holding

## What makes the estimands differ

Any difference between them is driven by two things

- Variation in treatment effects across covariate profiles
- Differences in the distribution of those profiles across treated and untreated
  units

## {.center background-color="#23373B"}

### **Interaction terms parameterize heterogeneity**

## A constant treatment effect

- A model like `y ~ a + l` says the conditional mean is `b0 + b1 * a + b2 * l`
- The contrast between `a = 1` and `a = 0` is `b1`, whatever the value of `l`

. . .

The treatment effect is assumed to be the same for every unit, regardless of
covariate values

## Letting the effect vary

- Suppose one variable in the confounder set modifies the treatment effect, and
  call it `z`
- `y ~ a * z + l` adds the product term:
  `b0 + b1 * a + b2 * z + b3 * a * z + b4 * l`

. . .

The conditional treatment effect becomes `b1 + b3 * z`, which depends on `z`

## When is the effect constant?

- If `b3` is zero, the treatment effect is constant
- Otherwise the treatment effect changes across values of `z`

. . .

The interaction term formalizes the claim that the effect of treatment differs
across levels of a covariate

## Interaction or separate models?

- An equivalent representation is to fit separate models within strata of `z`,
  which lets the treatment coefficient differ across groups
- The interaction model is often preferred because it estimates those
  subgroup-specific effects within a single framework, using all the data

## {.center background-color="#23373B"}

### **A simulation where we know the answer**

## Simulate the potential outcomes

```{r}
#| code-line-numbers: "|5|6|7-8|9|10"

set.seed(1000)
n <- 10000

sim_data <- tibble(
  l = rnorm(n),
  z = factor(rbinom(n, 1, 0.5), labels = c("low", "high")),
  y0 = 10 + 2 * l + 3 * (z == "high") + rnorm(n),
  y1 = y0 + 2 + 4 * (z == "high"),
  a = rbinom(n, 1, plogis(-1.25 + 0.5 * l + 2.5 * (z == "high"))),
  y = if_else(a == 1, y1, y0)
)
```

## What the simulation says

- `l` is a continuous confounder: it affects both the exposure and the outcome
- `z` is a confounder too, and it also modifies the effect: the individual
  effect is 2 in the `low` stratum and 6 in the `high` stratum

. . .

Both potential outcomes are simulated, so the answer is exact in this sample

## Who gets treated

```{r}
#| output-location: fragment

sim_data |>
  count(a, z) |>
  mutate(prop = n / sum(n), .by = a)
```

. . .

About three quarters of the treated are `high`, and about three quarters of the
untreated are `low`

## The truth in this simulation

```{r}
#| output-location: fragment
#| code-line-numbers: "|2|4-6"

sim_data |>
  mutate(effect = y1 - y0) |>
  summarize(
    ate = mean(effect),
    att = mean(effect[a == 1]),
    atc = mean(effect[a == 0])
  )
```

. . .

The ATT is larger than the ATC because the treated are mostly `high`, where the
effect is 6

## {.center background-color="#23373B"}

### **G-computation with an interaction**

## Fit the interaction model

```{r}
#| output-location: fragment

gcomp_model <- lm(y ~ a * z + l, data = sim_data)

tidy(gcomp_model)
```

## The coefficients are conditional

- `a` is the effect in the `low` stratum, and `a:zhigh` is how much larger the
  effect is in the `high` stratum
- Together they say the effect is about 2 and about 6

. . .

That is the conditional story, and averaging it over a population by eye is
another matter

## `avg_comparisons()` for the ATE

```{r}
#| output-location: fragment

avg_comparisons(gcomp_model, variables = "a")
```

## Targeting the ATT and the ATC

```{r}
#| eval: false
#| code-line-numbers: "|5,12"

# ATT: average over the exposed observations
avg_comparisons(
  gcomp_model,
  variables = "a",
  newdata = filter(sim_data, a == 1)
)

# ATC: average over the unexposed observations
avg_comparisons(
  gcomp_model,
  variables = "a",
  newdata = filter(sim_data, a == 0)
)
```

## Same model, three populations

```{r}
#| echo: false

bind_rows(
  ATE = tidy(avg_comparisons(gcomp_model, variables = "a")),
  ATT = tidy(avg_comparisons(
    gcomp_model,
    variables = "a",
    newdata = filter(sim_data, a == 1)
  )),
  ATC = tidy(avg_comparisons(
    gcomp_model,
    variables = "a",
    newdata = filter(sim_data, a == 0)
  )),
  .id = "estimand"
) |>
  select(estimand, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 2)))
```

Each estimate matches its simulated truth

## Averaging one response surface

Once interaction is introduced, these estimands can differ because they average
the same heterogeneous response surface over different covariate distributions

## Effects within levels of `z`

```{r}
#| output-location: fragment

avg_comparisons(gcomp_model, variables = "a", by = "z")
```

## Testing the difference

```{r}
#| output-location: fragment
#| code-line-numbers: "|5"

avg_comparisons(
  gcomp_model,
  variables = "a",
  by = "z",
  hypothesis = ~pairwise
)
```

. . .

The two strata differ by about 4, with a standard error of about 0.05

## The effect in each stratum

```{r}
#| output-location: slide

plot_comparisons(gcomp_model, variables = "a", by = "z") +
  expand_limits(y = 0) +
  labs(x = "modifier", y = "effect of a on y")
```

## Specify every interaction

Because g-computation focuses on modeling the *outcome*, we must specify every
interaction we think exists in order to estimate treatment effect heterogeneity

## Drop the interaction

```{r}
gcomp_model_add <- lm(y ~ a + z + l, data = sim_data)
```

```{r}
#| echo: false

bind_rows(
  ATE = tidy(avg_comparisons(gcomp_model_add, variables = "a")),
  ATT = tidy(avg_comparisons(
    gcomp_model_add,
    variables = "a",
    newdata = filter(sim_data, a == 1)
  )),
  ATC = tidy(avg_comparisons(
    gcomp_model_add,
    variables = "a",
    newdata = filter(sim_data, a == 0)
  )),
  .id = "estimand"
) |>
  select(estimand, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 2)))
```

Without an exposure-covariate interaction, the three coincide: the model assumes
a constant effect across covariate values

## {.center background-color="#23373B"}

### **The same estimands through weights**

## IPW targets the same estimands

- IPW does not require an explicit regression model for the outcome surface
- The weighting scheme determines which covariate distribution the heterogeneous
  unit-level effects are averaged over

. . .

IPW captures heterogeneity implicitly through the weighting and averaging
target, not by estimating subgroup-specific effects

## Fit the propensity score model

```{r}
#| code-line-numbers: "|1|3-4|6-8"

propensity_model <- glm(a ~ l + z, family = binomial(), data = sim_data)

sim_data_wts <- propensity_model |>
  augment(type.predict = "response", data = sim_data) |>
  mutate(
    w_ate = wt_ate(.fitted, a),
    w_att = wt_att(.fitted, a),
    w_atu = wt_atu(.fitted, a)
  )
```

The modifier is a confounder, so it belongs in the propensity score model

## Three weighted outcome models

```{r}
outcome_ate <- lm(y ~ a, data = sim_data_wts, weights = w_ate)
outcome_att <- lm(y ~ a, data = sim_data_wts, weights = w_att)
outcome_atu <- lm(y ~ a, data = sim_data_wts, weights = w_atu)
```

- Each model regresses the outcome on the exposure alone, and only the weights
  change
- `wt_atu()` builds the weights for the unexposed observations, the ATU or ATC

## Estimates through `ipw()`

```{r}
#| output-location: fragment

ipw(propensity_model, outcome_ate) |>
  tidy(conf.int = TRUE) |>
  select(term, estimate, std.error, conf.low, conf.high)
```

## The three weighted estimates

```{r}
#| echo: false

bind_rows(
  ATE = ipw(propensity_model, outcome_ate) |> tidy(conf.int = TRUE),
  ATT = ipw(propensity_model, outcome_att) |> tidy(conf.int = TRUE),
  ATC = ipw(propensity_model, outcome_atu) |> tidy(conf.int = TRUE),
  .id = "estimand"
) |>
  select(estimand, estimate, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 2)))
```

The weighted estimates recover the same truths as g-computation

## The weights carry the estimand

- The weighted outcome model has no covariates, so the differences among the
  three estimates flow entirely through the weights
- The modifier separates the exposure groups strongly here, so the spread
  between the ATT and the ATC is visible

## {.center background-color="#23373B"}

### **Effect modification with `ipw()`**

## The subgroup question with weights

```{r}
#| output-location: fragment
#| code-line-numbers: "|1|3"

outcome_model_int <- lm(y ~ a * z, data = sim_data_wts, weights = w_ate)

ipw(propensity_model, outcome_model_int, .by = z) |>
  tidy(conf.int = TRUE) |>
  select(group, estimate, std.error, conf.low, conf.high)
```

## {.center background-color="#23373B"}

### **Stabilizing the weights**

## Stabilized weights, revisited

- `stabilize = TRUE` puts the marginal probability of the observed exposure in
  the numerator, as it did when we first built ATE weights
- Under effect modification, the numerator should condition on the modifier
  instead

## Condition the numerator on `z`

```{r}
#| code-line-numbers: "|1-2|10"

numerator_model <- glm(a ~ z, family = binomial(), data = sim_data)
p_z <- fitted(numerator_model)

sim_data_wts <- sim_data_wts |>
  mutate(
    sw_ate = wt_ate(
      .fitted,
      a,
      stabilize = TRUE,
      stabilization_score = if_else(a == 1, p_z, 1 - p_z)
    )
  )
```

`fitted()` alone gives the probability of exposure for every unit, which is not
the probability of the exposure an unexposed unit took

## The rule for the numerator

- Conditioning the numerator on a variable that is also in the outcome model
  does not change the estimand
- Conditioning on a variable that is not in the outcome model does change it:
  the pseudo-population is one where that variable still predicts exposure

. . .

Only condition the numerator on variables that are already in the outcome model

## The weights tighten

```{r}
#| output-location: fragment

sim_data_wts |>
  select(w_ate, sw_ate) |>
  # the weights stack only as plain numbers
  mutate(across(everything(), as.numeric)) |>
  pivot_longer(everything(), names_to = "weights") |>
  summarize(sd = sd(value), max = max(value), .by = weights)
```

. . .

Conditioning the numerator on `z` cuts the spread of the weights and pulls in
the largest ones

## The same fit, stabilized

```{r}
#| output-location: fragment

outcome_model_stab <- lm(y ~ a * z, data = sim_data_wts, weights = sw_ate)

ipw(propensity_model, outcome_model_stab, .by = z) |>
  tidy(conf.int = TRUE) |>
  select(group, estimate, std.error, conf.low, conf.high)
```

## {.center background-color="#23373B"}

### **Three jobs for interaction terms**

## The same syntax, three jobs

- **Functional form**: an interaction between two covariates, in either model
- **Effect modification**: an interaction between the exposure and a covariate
- **Causal interaction**: an interaction between two exposures

## The same syntax, three jobs

The more exposures the term involves, the more causal the claim it makes

## Covariate by covariate

- In the diagnostics module we added `income * health` to the propensity score
  model because that product term was imbalanced
- Outcome models need the same care: g-computation averages predictions from
  the model we fit, so a term the conditional mean needs but the model lacks
  biases that average

## Covariate by covariate

Neither term involves the exposure, so neither says anything about the effect

## Exposure by covariate

- The interaction in this module's outcome model is between the exposure and
  `z`, so it says the effect itself differs across levels of `z`
- `z` is a covariate we condition on rather than something we set

. . .

The third job needs a second exposure rather than a covariate

## {.center background-color="#23373B"}

### **Effect modification versus causal interaction**

## Two different causal questions

- **Effect modification** asks whether the exposure's effect differs across
  levels of a pre-treatment variable
- **Causal interaction** asks whether the joint effect of changing both variables
  differs from what their separate effects predict

. . .

These are often used interchangeably, but they answer different questions

## Effect modification

- The comparison is `E[y(1) - y(0) | z = high]` against
  `E[y(1) - y(0) | z = low]`
- `z` is used to define subgroups: we ask how the effect of the exposure changes
  across observed strata of `z`, not what would happen if we intervened on it

## Causal interaction

- Both variables are exposures of causal interest
- The comparison is `E[y(a = 1, b = 1) - y(a = 0, b = 1)]` against
  `E[y(a = 1, b = 0) - y(a = 0, b = 0)]`

. . .

The effect of one exposure under two settings of the other

## Extra Magic Mornings by quarter

- Does an Extra Magic Morning change posted wait times differently in the fourth
  quarter than in the first?
- No one can assign a day to a calendar quarter, so the quarter defines
  subgroups rather than an intervention

. . .

That is an effect modification question

## Extra Magic Mornings and season

- Disney sets ticket season through pricing, so both the morning policy and the
  season are things the park can change
- Asking what would happen if we set both at once is a joint intervention

. . .

That is a causal interaction question

## The model does not decide

The same algebraic model can serve either purpose, but the interpretation
depends on the underlying question

## {.center background-color="#23373B"}

### **Intervening on two exposures at once**

## Two exposures, one confounder

```{r}
#| code-line-numbers: "|5|6-7|8"

set.seed(1)
n <- 10000

joint_data <- tibble(
  l = rnorm(n),
  a = rbinom(n, 1, plogis(-0.5 + 0.8 * l)),
  b = rbinom(n, 1, plogis(-0.3 + 0.9 * l)),
  y = 10 + 2 * l + 1 * a + 2 * b + 2 * a * b + rnorm(n)
)
```

`l` confounds both exposures, which are otherwise independent given `l`

## What this simulation says

- The four counterfactual means are 10, 11, 12, and 15
- The effect of `a` is 1 when `b` is 0 and 3 when `b` is 1
- The difference between those two effects, the interaction, is 2

## Cross the two exposures

```{r}
#| output-location: fragment

joint_data <- joint_data |>
  mutate(joint = joint_exposure(a = a, b = b))

levels(joint_data$joint)
```

. . .

The crossing is one categorical exposure with a cell per combination

## One model over the cells

```{r}
#| code-line-numbers: "|1|3-4"

ps_joint <- multinom(joint ~ l, data = joint_data, trace = FALSE)

joint_data <- joint_data |>
  mutate(w_joint = wt_ate(fitted(ps_joint), joint))
```

The categorical route takes the matrix of fitted probabilities, one column per
cell

## The weighted outcome model

```{r}
#| output-location: slide
#| code-line-numbers: "|1|3-5"

joint_model <- lm(y ~ joint, data = joint_data, weights = w_joint)

ipw(ps_joint, joint_model) |>
  tidy(conf.int = TRUE) |>
  select(contrast, group, estimate, conf.low, conf.high)
```

## Nine rows for a two-by-two

- Four **cell means**: the outcome had everyone been set to that combination
- Four **simple effects**: each exposure's effect within a fixed level of the
  other
- One **interaction**: the difference between the first exposure's two simple
  effects

## Beyond the reference cell

The simple effects include comparisons that reporting each cell against the
reference cell cannot express at all

## The interaction row

The interaction recovers the simulated 2, and on this scale it is the additive
interaction

## One system, again

- Every row is a parameter of the same stacked system, so the interaction
  covaries with the simple effects
- The surface reports the ATE alone, and it does not combine with `.by`:
  modification of a joint intervention is a three-way question

. . .

No odds ratio appears anywhere on this surface, for the same noncollapsibility
reason

## Two models instead of one

A joint intervention can also be weighted through one model per treatment, with
the product of their weights, which is what you want when the two treatments
call for different adjustment sets

## {.center background-color="#23373B"}

### **Interaction in a DAG**

## A conventional DAG

```{r}
#| output-location: slide

conventional_dag <- dagify(
  X ~ Z,
  Q ~ Z,
  Y ~ X + Q + Z,
  coords = time_ordered_coords(),
  exposure = "X",
  outcome = "Y"
)

ggdag(conventional_dag) + theme_dag()
```

## Interaction is not in the graph

A conventional DAG shows that `X`, `Q`, and `Z` all affect `Y`, but not whether
`X` and `Q` interact

. . .

The interaction, if present, is implicit in the functional form of how they
jointly affect `Y`

## An explicit interaction node

```{r}
#| output-location: slide
#| code-line-numbers: "|5|6"

explicit_interaction_dag <- dagify(
  X ~ Z,
  Q ~ Z,
  Y ~ X + Q + Z + XxQ,
  XxQ ~ X + Q,
  coords = time_ordered_coords(),
  exposure = "X",
  outcome = "Y"
)

ggdag(explicit_interaction_dag) + theme_dag()
```

## Making the interaction visible

Arrows run from `X` and `Q` into `XxQ`, and from `XxQ` into `Y`, which can help
identify adjustment sets for estimating interaction effects

::: tiny
Attia J, Holliday E, Oldmeadow C (2022). A proposal for capturing interaction
and effect modification using DAGs. International Journal of Epidemiology.
doi:10.1093/ije/dyac126
:::

## *Your Turn 1* (**`14-interaction-exercises.qmd`**) {.small}

### Fit an outcome model for posted wait times that lets the effect of an Extra Magic Morning vary by ticket season

### Read the conditional effects off the coefficients

`r countdown::countdown(minutes = 5)`

## *Your Turn 2*

### Use `avg_comparisons()` to get the ATE, then the ATT and the ATC with `newdata`

### Compare the three estimates and explain why they differ

### *Stretch*: refit without the interaction and watch the three collapse

`r countdown::countdown(minutes = 6)`

## *Your Turn 3*

### Ask whether the effect of an Extra Magic Morning is modified by calendar quarter

### Fit the propensity score model, build ATE weights, and pass a weighted outcome model with the interaction to `ipw()` with `.by`

`r countdown::countdown(minutes = 8)`

## *Your Turn 4* {.small}

### Ask the joint question: what if the park set both the Extra Magic Morning policy and the ticket season?

### Cross the two with `joint_exposure()`, fit one propensity model over the cells, and read the cell means, simple effects, and interaction

### Finish early? Try the *stretch goals*

`r countdown::countdown(minutes = 10)`
