---
title: "Bonus: Missing Data"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

## {background-color="#23373B"}

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)
library(tidyverse)
library(broom)
library(propensity)
library(causalworkshop)
library(touringplans)
# mice masks tidyr::complete(), so tidyr::complete() calls must be namespaced
library(mice)
library(ggmice)
library(ggdag)
library(patchwork)

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)
```

- `lm()` and `glm()` drop rows with missing values silently
- Every model we have fit in this course has been a complete-case analysis,
  whether we said so or not

## We have said this before

- Our own exercises say: "We also have a lot of missingness for
  `wait_minutes_actual_avg`, so we'll drop unobserved values for now"
- We dropped the unobserved values, and we postponed the problem

. . .

This module is where we come back for it

## How much is missing?

```{r}
#| output-location: fragment

seven_dwarfs_train_2018 |>
  filter(wait_hour == 9) |>
  summarize(
    days = n(),
    missing_actual = sum(is.na(wait_minutes_actual_avg)),
    prop_missing = mean(is.na(wait_minutes_actual_avg))
  )
```

. . .

At the 9 AM hour, 207 of 354 days are missing the actual wait time; across the
full data, 88.7% of actual wait times are missing

## The plan for this module

- The malaria data has no missing values, so we will remove some on purpose
- Then we judge every fix against the answer we already know

. . .

The benchmark: the full-data IPW estimate of the effect of net use on malaria
risk, -12.5 (from the outcome model module)

## {.center background-color="#23373B"}

### **The missingness indicator**

## Two nodes for one variable

- We draw two nodes for a given variable: one represents the true values, and
  the other represents a *missingness indicator*, whether we observed the value
  or not
- We are always conditioning on the data we actually have

. . .

With missingness, that usually means conditioning on *complete* observations:
the subset where every variable we need is observed

## Missingness has its own causes

```{r}
#| echo: false

malaria_labels <- c(
  malaria_risk = "Risk of malaria",
  malaria_missing = "Missingness\nin malaria risk",
  net = "Mosquito net",
  unknown = "Unknown\nmechanism"
)

missing_malaria <- dagify(
  malaria_risk ~ net,
  malaria_missing ~ net + unknown,
  coords = list(
    x = c(net = 0, unknown = 0, malaria_risk = 2, malaria_missing = 2),
    y = c(net = 1, unknown = -1, malaria_risk = 1, malaria_missing = -1)
  ),
  labels = malaria_labels,
  exposure = "net",
  outcome = "malaria_risk"
)

missing_malaria |>
  ggdag(use_text = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  expand_plot(expansion(0.2), expansion(0.2))
```

## Reading this DAG

- The missingness of malaria risk is related to net use and to an unknown
  mechanism
- The unknown mechanism is random, but the mechanism related to net use is not,
  since net use is the exposure

## {.center background-color="#23373B"}

### **When complete cases are enough**

## Conditioning on the indicator

```{r}
#| echo: false

missing_malaria |>
  ggdag_adjust("malaria_missing", use_text = FALSE, collider_lines = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot(expansion(0.2), expansion(0.2)) +
  ggokabeito::scale_color_okabe_ito()
```

## A collider that opens no backdoor

- The missingness indicator is a collider, and conditioning on it is exactly
  what complete-case analysis does
- In this DAG, though, conditioning on it does not open a backdoor path between
  net use and malaria risk
- It does bias the relationship between the unknown mechanism and net use, but
  we do not care about that relationship

## What we keep and what we lose

- The causal effect is *recoverable*: we can still estimate it without bias
  with complete-case analysis
- Because we only use complete observations, we have a smaller sample size and
  thus reduced precision

. . .

We are unable to correctly estimate the *mean* of malaria risk because values
are systematically missing by net use

## {.center background-color="#23373B"}

### **When complete cases are not enough**

## Missingness from the value itself

- Is it feasible that the missingness of malaria risk is related to its own
  values?
- It could be that a person's malaria risk is so severe that the measurement
  never happens

. . .

The values of malaria risk themselves influence whether we managed to measure
them

## An arrow from the outcome

```{r}
#| echo: false

missing_malaria_own <- dagify(
  malaria_risk ~ net,
  malaria_missing ~ net + unknown + malaria_risk,
  coords = list(
    x = c(net = 0, unknown = 0, malaria_risk = 2, malaria_missing = 2),
    y = c(net = 1, unknown = -1, malaria_risk = 1, malaria_missing = -1)
  ),
  labels = malaria_labels,
  exposure = "net",
  outcome = "malaria_risk"
)

missing_malaria_own |>
  ggdag(use_text = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  expand_plot(expansion(0.2), expansion(0.2))
```

## Conditioning now opens a backdoor

```{r}
#| echo: false

missing_malaria_own |>
  ggdag_adjust("malaria_missing", use_text = FALSE, collider_lines = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot(expansion(0.2), expansion(0.2)) +
  ggokabeito::scale_color_okabe_ito()
```

## No way to close it

- Whether malaria risk was measured is now a descendant of both the exposure
  and the outcome
- Conditioning on the indicator opens a backdoor path, creating
  nonexchangeability

. . .

There is no way to close the backdoor path opened by conditioning on
missingness

## More than backdoor paths

- What effects we can recover is more complicated than determining backdoor
  paths alone
- When there are no backdoor paths, we can calculate the causal effect, but we
  may not be able to calculate other statistics, like the mean of the exposure
  or outcome
- When conditioning on missingness opens a backdoor path, sometimes we can
  close it, and sometimes we cannot

## {.center background-color="#23373B"}

### **Five structures of missingness**

## Five DAGs, four letters

- Consider five DAGs, where `a` is malaria risk, `n` is net use, `u` is
  unknown, and `m` is missing
- Each represents a simple but differing structure of missingness

. . .

In DAGs 1-3, malaria risk has missingness; in DAGs 4-5, net use does

## Five structures

```{r}
#| echo: false

define_dag <- function(..., tag, title) {
  dagify(
    ...,
    coords = time_ordered_coords(),
    exposure = "n",
    outcome = "a"
  ) |>
    ggdag(size = 0.7) +
    labs(title = paste0(tag, ": ", title)) +
    theme_dag() +
    theme(plot.title = element_text(size = 12)) +
    expand_plot(expansion(c(0.4, 0.4)), expansion(c(0.4, 0.4)))
}

dag_1 <- define_dag(
  a ~ n,
  m ~ u,
  tag = "1",
  title = "malaria risk is missing"
)

dag_2 <- define_dag(
  a ~ n,
  m ~ u + n,
  tag = "2",
  title = "malaria risk is missing"
)

dag_3 <- define_dag(
  a ~ n,
  m ~ u + a,
  tag = "3",
  title = "malaria risk is missing"
)

dag_4 <- define_dag(
  a ~ n,
  m ~ u + n,
  tag = "4",
  title = "net use is missing"
)

dag_5 <- define_dag(
  a ~ n,
  m ~ u + a,
  tag = "5",
  title = "net use is missing"
)

(dag_1 + dag_2 + dag_3) / (plot_spacer() + dag_4 + dag_5)
```

## Impose the five structures on the data

```{r}
set.seed(123)
n <- nrow(net_data)
m1 <- as.logical(rbinom(n, 1, 0.35))
m2 <- as.logical(
  if_else(net_data$net, rbinom(n, 1, 0.9), rbinom(n, 1, 0.15))
)
# the threshold 38 sits near the median of malaria risk
m3 <- as.logical(
  if_else(net_data$malaria_risk > 38, rbinom(n, 1, 0.9), rbinom(n, 1, 0.05))
)
# the same structures, but net use gets the resulting missingness
m4 <- m2
m5 <- m3
```

. . .

For each structure, we blank out the chosen column, then estimate the
proportion using nets, the mean of malaria risk, and the IPW effect

## What each structure leaves standing

```{r}
#| echo: false

fit_ipw_net <- function(.data) {
  propensity_model <- glm(
    net ~ income + health + temperature,
    family = binomial(),
    data = .data
  )
  weighted_data <- propensity_model |>
    augment(type.predict = "response", data = .data) |>
    mutate(wts = wt_ate(.fitted, net, exposure_type = "binary"))
  ipw_model <- lm(malaria_risk ~ net, data = weighted_data, weights = wts)
  ipw(propensity_model, ipw_model)
}

fit_stats <- function(dag, miss, missing_for) {
  d <- net_data
  if (missing_for == "outcome") {
    d$malaria_risk[miss] <- NA
  } else {
    d$net[miss] <- NA
  }

  t_risk <- t.test(d$malaria_risk)
  p_net <- prop.test(sum(d$net, na.rm = TRUE), sum(!is.na(d$net)))
  # ipw() requires both models to see the same rows, so filter to complete
  # cases before fitting instead of letting lm() drop rows
  complete_cases <- d |> filter(!is.na(malaria_risk), !is.na(net))
  effect <- tidy(fit_ipw_net(complete_cases), conf.int = TRUE)

  tibble(
    dag = dag,
    prop_net_estimate = unname(p_net$estimate),
    prop_net_lower = p_net$conf.int[[1]],
    prop_net_upper = p_net$conf.int[[2]],
    mean_risk_estimate = unname(t_risk$estimate),
    mean_risk_lower = t_risk$conf.int[[1]],
    mean_risk_upper = t_risk$conf.int[[2]],
    effect_estimate = effect$estimate,
    effect_lower = effect$conf.low,
    effect_upper = effect$conf.high
  ) |>
    pivot_longer(
      cols = -dag,
      names_to = c("stat", ".value"),
      names_pattern = "^(.*)_(estimate|lower|upper)$"
    )
}

dag_stats <- bind_rows(
  fit_stats("No missingness", rep(FALSE, n), "outcome"),
  fit_stats("DAG 1", m1, "outcome"),
  fit_stats("DAG 2", m2, "outcome"),
  fit_stats("DAG 3", m3, "outcome"),
  fit_stats("DAG 4", m4, "exposure"),
  fit_stats("DAG 5", m5, "exposure")
)

dag_stats |>
  mutate(
    true_value = if_else(
      dag == "No missingness",
      "True value",
      "Observed value"
    ),
    dag = factor(dag, levels = c(paste("DAG", 5:1), "No missingness")),
    stat = factor(
      stat,
      levels = c("prop_net", "mean_risk", "effect"),
      labels = c(
        "Proportion using nets",
        "Mean of malaria risk",
        "Causal effect"
      )
    )
  ) |>
  ggplot(aes(color = true_value)) +
  geom_point(aes(estimate, dag)) +
  geom_segment(aes(
    x = lower,
    xend = upper,
    y = dag,
    yend = dag,
    group = stat
  )) +
  facet_wrap(~stat, scales = "free_x") +
  labs(y = NULL, color = NULL) +
  ggokabeito::scale_color_okabe_ito()
```

## What each structure costs

- With no missingness, we can estimate all three quantities
- DAG 1 loses only precision
- DAG 2 loses the mean of malaria risk
- DAG 3 loses the mean of malaria risk and the causal effect
- DAG 4 loses the proportion using nets
- DAG 5 loses everything but the mean of malaria risk

. . .

The causal effect is recoverable in structures 1, 2, and 4, and lost in 3 and 5

## A comprehensive map

See Moreno-Betancur et al. (2018) for a comprehensive overview of what effects
are recoverable from different structures of missingness

::: tiny
Moreno-Betancur M, Lee KJ, Leacy FP, White IR, Simpson JA, Carlin JB (2018).
Canonical causal diagrams to guide the treatment of missing data in
epidemiologic studies. American Journal of Epidemiology.
doi:10.1093/aje/kwy173
:::

## {.center background-color="#23373B"}

### **MCAR, MAR, and MNAR**

## Three terms you will see

In the grand tradition of statisticians being bad at naming things, you will
also commonly see missingness discussed in terms of *missing completely at
random*, *missing at random*, and *missing not at random*

. . .

We can explain these ideas by the causal structure of missingness and the
availability of variables and values related to that structure in your data

## Missing completely at random

- **MCAR**: there are missing values, but the causes of missingness are such
  that the missingness process is unrelated to the causal structure of your
  question
- In other words, the only problem with missingness is a reduction in sample
  size

. . .

This is structure 1, where the only cause of missingness is the unknown
mechanism

## Missing at random

- **MAR**: the causes of missingness are related to the causal structure of
  the research problem, but missingness only depends on the variables and
  values in the data that we have actually observed

. . .

This is structure 2, where the missingness of malaria risk depends on net use,
which we always observe

## Missing not at random

- **MNAR**: the causes of missingness are related to the causal structure of
  the research problem, but the process is related to values we are missing
- A classic example is when a variable's missingness is impacted by itself:
  higher values of `x` are more likely to be missing in `x`, as in structure 3

. . .

We do not have that information because, by definition, it is missing

## Names are not instructions

- These terms do not always tell you what to do next
- We will avoid them in favor of explicitly describing the missingness
  generation process

. . .

Next: impose one realistic mechanism on the malaria data and try to repair it
