---
title: "Bonus: Missing Data"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

## {background-color="#23373B"}

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)
library(tidyverse)
library(broom)
library(propensity)
library(causalworkshop)
library(touringplans)
# mice masks tidyr::complete(), so tidyr::complete() calls must be namespaced
library(mice)
library(ggmice)
library(ggdag)
library(patchwork)

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)
```

- `lm()` and `glm()` drop rows with missing values silently
- Every model we have fit in this course has been a complete-case analysis,
  whether we said so or not

## We have said this before

- Our own exercises say: "We also have a lot of missingness for
  `wait_minutes_actual_avg`, so we'll drop unobserved values for now"
- We dropped the unobserved values, and we postponed the problem

. . .

This module is where we come back for it

## How much is missing?

```{r}
#| output-location: fragment

seven_dwarfs_train_2018 |>
  filter(wait_hour == 9) |>
  summarize(
    days = n(),
    missing_actual = sum(is.na(wait_minutes_actual_avg)),
    prop_missing = mean(is.na(wait_minutes_actual_avg))
  )
```

. . .

At the 9 AM hour, 207 of 354 days are missing the actual wait time; across the
full data, 88.7% of actual wait times are missing

## The plan for this module

- The malaria data has no missing values, so we will remove some on purpose
- Then we judge every fix against the answer we already know

. . .

The benchmark: the full-data IPW estimate of the effect of net use on malaria
risk, -12.5 (from the outcome model module)

## {.center background-color="#23373B"}

### **The missingness indicator**

## Two nodes for one variable

- We draw two nodes for a given variable: one represents the true values, and
  the other represents a *missingness indicator*, whether we observed the value
  or not
- We are always conditioning on the data we actually have

. . .

With missingness, that usually means conditioning on *complete* observations:
the subset where every variable we need is observed

## Missingness has its own causes

```{r}
#| echo: false

malaria_labels <- c(
  malaria_risk = "Risk of malaria",
  malaria_missing = "Missingness\nin malaria risk",
  net = "Mosquito net",
  unknown = "Unknown\nmechanism"
)

missing_malaria <- dagify(
  malaria_risk ~ net,
  malaria_missing ~ net + unknown,
  coords = list(
    x = c(net = 0, unknown = 0, malaria_risk = 2, malaria_missing = 2),
    y = c(net = 1, unknown = -1, malaria_risk = 1, malaria_missing = -1)
  ),
  labels = malaria_labels,
  exposure = "net",
  outcome = "malaria_risk"
)

missing_malaria |>
  ggdag(use_text = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  expand_plot(expansion(0.2), expansion(0.2))
```

## Reading this DAG

- The missingness of malaria risk is related to net use and to an unknown
  mechanism
- The unknown mechanism is random, but the mechanism related to net use is not,
  since net use is the exposure

## {.center background-color="#23373B"}

### **When complete cases are enough**

## Conditioning on the indicator

```{r}
#| echo: false

missing_malaria |>
  ggdag_adjust("malaria_missing", use_text = FALSE, collider_lines = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot(expansion(0.2), expansion(0.2)) +
  ggokabeito::scale_color_okabe_ito()
```

## A collider that opens no backdoor

- The missingness indicator is a collider, and conditioning on it is exactly
  what complete-case analysis does
- In this DAG, though, conditioning on it does not open a backdoor path between
  net use and malaria risk
- It does bias the relationship between the unknown mechanism and net use, but
  we do not care about that relationship

## What we keep and what we lose

- The causal effect is *recoverable*: we can still estimate it without bias
  with complete-case analysis
- Because we only use complete observations, we have a smaller sample size and
  thus reduced precision

. . .

We are unable to correctly estimate the *mean* of malaria risk because values
are systematically missing by net use

## {.center background-color="#23373B"}

### **When complete cases are not enough**

## Missingness from the value itself

- Is it feasible that the missingness of malaria risk is related to its own
  values?
- It could be that a person's malaria risk is so severe that the measurement
  never happens

. . .

The values of malaria risk themselves influence whether we managed to measure
them

## An arrow from the outcome

```{r}
#| echo: false

missing_malaria_own <- dagify(
  malaria_risk ~ net,
  malaria_missing ~ net + unknown + malaria_risk,
  coords = list(
    x = c(net = 0, unknown = 0, malaria_risk = 2, malaria_missing = 2),
    y = c(net = 1, unknown = -1, malaria_risk = 1, malaria_missing = -1)
  ),
  labels = malaria_labels,
  exposure = "net",
  outcome = "malaria_risk"
)

missing_malaria_own |>
  ggdag(use_text = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  expand_plot(expansion(0.2), expansion(0.2))
```

## Conditioning now opens a backdoor

```{r}
#| echo: false

missing_malaria_own |>
  ggdag_adjust("malaria_missing", use_text = FALSE, collider_lines = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot(expansion(0.2), expansion(0.2)) +
  ggokabeito::scale_color_okabe_ito()
```

## No way to close it

- Whether malaria risk was measured is now a descendant of both the exposure
  and the outcome
- Conditioning on the indicator opens a backdoor path, creating
  nonexchangeability

. . .

There is no way to close the backdoor path opened by conditioning on
missingness

## More than backdoor paths

- What effects we can recover is more complicated than determining backdoor
  paths alone
- When there are no backdoor paths, we can calculate the causal effect, but we
  may not be able to calculate other statistics, like the mean of the exposure
  or outcome
- When conditioning on missingness opens a backdoor path, sometimes we can
  close it, and sometimes we cannot

## {.center background-color="#23373B"}

### **Five structures of missingness**

## Five DAGs, four letters

- Consider five DAGs, where `a` is malaria risk, `n` is net use, `u` is
  unknown, and `m` is missing
- Each represents a simple but differing structure of missingness

. . .

In DAGs 1-3, malaria risk has missingness; in DAGs 4-5, net use does

## Five structures

```{r}
#| echo: false

define_dag <- function(..., tag, title) {
  dagify(
    ...,
    coords = time_ordered_coords(),
    exposure = "n",
    outcome = "a"
  ) |>
    ggdag(size = 0.7) +
    labs(title = paste0(tag, ": ", title)) +
    theme_dag() +
    theme(plot.title = element_text(size = 12)) +
    expand_plot(expansion(c(0.4, 0.4)), expansion(c(0.4, 0.4)))
}

dag_1 <- define_dag(
  a ~ n,
  m ~ u,
  tag = "1",
  title = "malaria risk is missing"
)

dag_2 <- define_dag(
  a ~ n,
  m ~ u + n,
  tag = "2",
  title = "malaria risk is missing"
)

dag_3 <- define_dag(
  a ~ n,
  m ~ u + a,
  tag = "3",
  title = "malaria risk is missing"
)

dag_4 <- define_dag(
  a ~ n,
  m ~ u + n,
  tag = "4",
  title = "net use is missing"
)

dag_5 <- define_dag(
  a ~ n,
  m ~ u + a,
  tag = "5",
  title = "net use is missing"
)

(dag_1 + dag_2 + dag_3) / (plot_spacer() + dag_4 + dag_5)
```

## Impose the five structures on the data

```{r}
set.seed(123)
n <- nrow(net_data)
m1 <- as.logical(rbinom(n, 1, 0.35))
m2 <- as.logical(
  if_else(net_data$net, rbinom(n, 1, 0.9), rbinom(n, 1, 0.15))
)
# the threshold 38 sits near the median of malaria risk
m3 <- as.logical(
  if_else(net_data$malaria_risk > 38, rbinom(n, 1, 0.9), rbinom(n, 1, 0.05))
)
# the same structures, but net use gets the resulting missingness
m4 <- m2
m5 <- m3
```

. . .

For each structure, we blank out the chosen column, then estimate the
proportion using nets, the mean of malaria risk, and the IPW effect

## What each structure leaves standing

```{r}
#| echo: false

# fit_ipw_net() is defined again on the visible pipeline slide below; keep
# the two definitions in sync
fit_ipw_net <- function(.data) {
  propensity_model <- glm(
    net ~ income + health + temperature,
    family = binomial(),
    data = .data
  )
  weighted_data <- propensity_model |>
    augment(type.predict = "response", data = .data) |>
    mutate(wts = wt_ate(.fitted, net, exposure_type = "binary"))
  ipw_model <- lm(malaria_risk ~ net, data = weighted_data, weights = wts)
  ipw(propensity_model, ipw_model)
}

fit_stats <- function(dag, miss, missing_for) {
  d <- net_data
  if (missing_for == "outcome") {
    d$malaria_risk[miss] <- NA
  } else {
    d$net[miss] <- NA
  }

  t_risk <- t.test(d$malaria_risk)
  p_net <- prop.test(sum(d$net, na.rm = TRUE), sum(!is.na(d$net)))
  # ipw() requires both models to see the same rows, so filter to complete
  # cases before fitting instead of letting lm() drop rows
  complete_cases <- d |> filter(!is.na(malaria_risk), !is.na(net))
  effect <- tidy(fit_ipw_net(complete_cases), conf.int = TRUE)

  tibble(
    dag = dag,
    prop_net_estimate = unname(p_net$estimate),
    prop_net_lower = p_net$conf.int[[1]],
    prop_net_upper = p_net$conf.int[[2]],
    mean_risk_estimate = unname(t_risk$estimate),
    mean_risk_lower = t_risk$conf.int[[1]],
    mean_risk_upper = t_risk$conf.int[[2]],
    effect_estimate = effect$estimate,
    effect_lower = effect$conf.low,
    effect_upper = effect$conf.high
  ) |>
    pivot_longer(
      cols = -dag,
      names_to = c("stat", ".value"),
      names_pattern = "^(.*)_(estimate|lower|upper)$"
    )
}

dag_stats <- bind_rows(
  fit_stats("No missingness", rep(FALSE, n), "outcome"),
  fit_stats("DAG 1", m1, "outcome"),
  fit_stats("DAG 2", m2, "outcome"),
  fit_stats("DAG 3", m3, "outcome"),
  fit_stats("DAG 4", m4, "exposure"),
  fit_stats("DAG 5", m5, "exposure")
)

dag_stats |>
  mutate(
    true_value = if_else(
      dag == "No missingness",
      "True value",
      "Observed value"
    ),
    dag = factor(dag, levels = c(paste("DAG", 5:1), "No missingness")),
    stat = factor(
      stat,
      levels = c("prop_net", "mean_risk", "effect"),
      labels = c(
        "Proportion using nets",
        "Mean of malaria risk",
        "Causal effect"
      )
    )
  ) |>
  ggplot(aes(color = true_value)) +
  geom_point(aes(estimate, dag)) +
  geom_segment(aes(
    x = lower,
    xend = upper,
    y = dag,
    yend = dag,
    group = stat
  )) +
  facet_wrap(~stat, scales = "free_x") +
  labs(y = NULL, color = NULL) +
  ggokabeito::scale_color_okabe_ito()
```

## What each structure costs

- With no missingness, we can estimate all three quantities
- DAG 1 loses only precision
- DAG 2 loses the mean of malaria risk
- DAG 3 loses the mean of malaria risk and the causal effect
- DAG 4 loses the proportion using nets
- DAG 5 loses everything but the mean of malaria risk

. . .

The causal effect is recoverable in structures 1, 2, and 4, and lost in 3 and 5

## A comprehensive map

See Moreno-Betancur et al. (2018) for a comprehensive overview of what effects
are recoverable from different structures of missingness

::: tiny
Moreno-Betancur M, Lee KJ, Leacy FP, White IR, Simpson JA, Carlin JB (2018).
Canonical causal diagrams to guide the treatment of missing data in
epidemiologic studies. American Journal of Epidemiology.
doi:10.1093/aje/kwy173
:::

## {.center background-color="#23373B"}

### **MCAR, MAR, and MNAR**

## Three terms you will see

In the grand tradition of statisticians being bad at naming things, you will
also commonly see missingness discussed in terms of *missing completely at
random*, *missing at random*, and *missing not at random*

. . .

We can explain these ideas by the causal structure of missingness and the
availability of variables and values related to that structure in your data

## Missing completely at random

- **MCAR**: there are missing values, but the causes of missingness are such
  that the missingness process is unrelated to the causal structure of your
  question
- In other words, the only problem with missingness is a reduction in sample
  size

. . .

This is structure 1, where the only cause of missingness is the unknown
mechanism

## Missing at random

- **MAR**: the causes of missingness are related to the causal structure of
  the research problem, but missingness only depends on the variables and
  values in the data that we have actually observed

. . .

This is structure 2, where the missingness of malaria risk depends on net use,
which we always observe

## Missing not at random

- **MNAR**: the causes of missingness are related to the causal structure of
  the research problem, but the process is related to values we are missing
- A classic example is when a variable's missingness is impacted by itself:
  higher values of `x` are more likely to be missing in `x`, as in structure 3

. . .

We do not have that information because, by definition, it is missing

## Names are not instructions

- These terms do not always tell you what to do next
- We will avoid them in favor of explicitly describing the missingness
  generation process

. . .

Next: impose one realistic mechanism on the malaria data and try to repair it

## {.center background-color="#23373B"}

### **Regression imputation**

## One realistic mechanism

```{r}
p_missing <- plogis(
  -0.8 + 1.4 * net_data$net + 0.3 * as.numeric(scale(net_data$income))
)
set.seed(88)
missing_mar <- rbinom(n, 1, p_missing) == 1

net_data_mar <- net_data |>
  select(net, income, health, temperature, malaria_risk) |>
  mutate(malaria_risk = if_else(missing_mar, NA_real_, malaria_risk))
```

. . .

Malaria risk now goes missing more often for net users and for higher-income
households: about 40% of the values are gone, but only for reasons we can see
in the data

## The pipeline, wrapped in a function

```{r}
fit_ipw_net <- function(.data) {
  propensity_model <- glm(
    net ~ income + health + temperature,
    family = binomial(),
    data = .data
  )
  weighted_data <- propensity_model |>
    augment(type.predict = "response", data = .data) |>
    mutate(wts = wt_ate(.fitted, net, exposure_type = "binary"))
  ipw_model <- lm(malaria_risk ~ net, data = weighted_data, weights = wts)
  ipw(propensity_model, ipw_model)
}
```

. . .

The pipeline from the outcome model module: propensity score model, ATE
weights, weighted outcome model, `ipw()`

## Complete cases first

```{r}
#| output-location: fragment

net_data_complete <- net_data_mar |> drop_na()

mar_complete_case <- fit_ipw_net(net_data_complete) |>
  tidy(conf.int = TRUE)

mar_complete_case
```

. . .

-12.8 (95% CI -14.3, -11.2): the effect survives with a wider interval, just
as the missingness indicator DAG promised

## Say the choice out loud

- We filtered to the complete cases ourselves before fitting either model
- Had we let `lm()` drop the missing rows silently, `ipw()` would error: the
  propensity score model would see all 1752 rows and the outcome model only
  the 1047 complete ones
- Complete-case analysis is an analytic choice, and the tooling makes us state
  it explicitly

. . .

The effect is recoverable here, but the mean of malaria risk is not: the
complete cases average 42.6 against the full-data 39.7

## Predict what is missing

- The analysis variables predict malaria risk well, so we can use them to fill
  in the values we failed to measure
- Fit a model for malaria risk on the complete cases, then plug its prediction
  in wherever the value is missing
- When fitting imputation models, include every variable that appears in your
  later models, in the same form it takes there

## Regression imputation

```{r}
imputation_model <- lm(
  malaria_risk ~ net + income + health + temperature,
  data = net_data_mar
)

net_data_imputed <- imputation_model |>
  augment(newdata = net_data_mar) |>
  mutate(malaria_risk = coalesce(malaria_risk, .fitted))
```

. . .

`coalesce()` keeps the observed value where we have one and uses the model's
prediction where we do not

## The point estimate is fine

```{r}
#| output-location: fragment

mar_regression <- fit_ipw_net(net_data_imputed) |>
  tidy(conf.int = TRUE)

mar_regression
```

. . .

-12.5 (95% CI -13.1, -11.8): right on the benchmark

## One guess, treated as truth

- Each of the 705 missing values got a single predicted value, and the outcome
  model then treated it like real data
- The standard error is 0.34, *narrower* than the 0.44 we had with no missing
  data at all
- Plugging in guesses cannot make us more certain than measuring everyone
- If you use this approach, include the fitting of the imputation model in
  your bootstrap so the standard errors reflect its uncertainty

. . .

Deterministic imputation understates uncertainty

## {.center background-color="#23373B"}

### **Multiple imputation**

## One guess is not enough

- Regression imputation relies on a single imputed value for each missing
  observation
- Done honestly, with the imputation model inside the bootstrap, this plug-in
  approach is generally inefficient: wide intervals that reflect how much
  rests on a single guess

. . .

Multiple imputation instead draws predicted values from a distribution,
capturing the uncertainty inherent in the missing data

## Many completed datasets

- We generate several completed datasets whose imputed values differ from draw
  to draw
- The values we observed are identical in every dataset; the values we imputed
  vary
- That variation between datasets is the imputation uncertainty, made visible

## Two rules for causal imputation models

1. The imputation model must include *every variable* in the outcome model and
   the propensity score model: the confounders, the exposure, and the outcome
   itself
2. Mirror the same functional forms you will use downstream: if a variable
   gets a spline or an interaction later, it needs one in the imputation model
   too

## The recipe

1. Impute multiple completed datasets with a model that includes the outcome
   and all covariates
2. Estimate the treatment effect within each dataset: propensity score model,
   weights, weighted outcome model
3. Pool the estimates with Rubin's rules to get one effect and valid standard
   errors

. . .

Rule of thumb: the number of imputations should roughly match the percentage
of incomplete cases

## Look at the pattern first

`plot_pattern()` shows which variables are missing and in what combinations:
here, one block of 705 rows missing malaria risk

```{r}
#| output-location: slide

plot_pattern(net_data_mar)
```

## The imputation model needs the interactions

- In the malaria data, the effect of net use on malaria risk varies with
  income
- An imputation model with only main effects would fill in values from an
  effect that is too weak, and the pooled estimate would inherit that bias
- This is rule 2 in action: the marginal effect we estimate averages over that
  heterogeneity, so the imputation model has to represent it

## Build the imputation frame

```{r}
net_data_to_impute <- net_data_mar |>
  mutate(
    net_income = as.numeric(net) * income,
    net_health = as.numeric(net) * health,
    net_temperature = as.numeric(net) * temperature
  )
```

. . .

The product columns are complete, so `mice()` never imputes them; they serve
as predictors in the imputation model for malaria risk

## Impute with mice()

```{r}
imp <- mice(
  net_data_to_impute,
  m = 40,
  method = "pmm",
  seed = 20260814,
  print = FALSE
)
```

. . .

- Predictive mean matching fills each missing value with an observed value
  borrowed from a case whose prediction is close
- About 40% of the cases are incomplete, so the rule of thumb says about 40
  imputations; we use `m = 40`

## Observed versus imputed

The imputed values should look like data, and their distribution should make
sense given the mechanism

```{r}
#| output-location: slide

ggmice(imp, aes(x = malaria_risk, group = .imp)) +
  geom_density()
```

## Reading the densities

- The observed values (blue) center higher than the imputed values (red)
- That is what the mechanism implies: net users are missing malaria risk more
  often, and net users have lower risk
- The imputation model can see this because net use is one of its predictors

## Did the algorithm converge?

- `mice()` imputes each variable with missingness iteratively, conditional on
  the others
- The trace plot shows the mean and spread of the imputed values across
  iterations; look for well-mixed lines with no trend

```{r}
#| output-location: slide

plot_trace(imp, "malaria_risk")
```

## Fit within each imputation

```{r}
fits <- with(imp, {
  ps <- glm(net ~ income + health + temperature, family = binomial())
  w <- wt_ate(fitted(ps), net, exposure_type = "binary")
  om <- lm(malaria_risk ~ net, weights = w)
  ipw(ps, om)
})
```

. . .

`with()` runs the block once per completed dataset: forty propensity score
models, forty sets of weights, forty outcome models, forty `ipw()` results

## Refit the propensity model every time

- The whole analysis is rebuilt inside each imputation, including the
  propensity score model
- Averaging the propensity scores across imputations and weighting with one
  averaged set hides the propensity model's estimation uncertainty from the
  pooling step
- Here only the outcome is imputed, so each refit returns the same model; when
  confounders or the exposure have missingness, the refits differ, and that
  variation belongs in the pooled standard error

. . .

The propensity score model is part of the analysis, so it belongs inside the
imputation loop

## What pooling means

```{r}
#| output-location: fragment

estimates <- map_dfr(fits$analyses, tidy)

estimates |>
  summarize(
    pooled = mean(estimate),
    within_var = mean(std.error^2),
    between_var = var(estimate),
    pooled_se = sqrt(within_var + (1 + 1 / n()) * between_var)
  )
```

. . .

The pooled estimate is the average across imputations; its variance is the
average within-imputation variance plus the between-imputation variance,
inflated by `1 + 1/m`

## pool_ipw() does it correctly

```{r}
#| output-location: fragment

mar_mi <- pool_ipw(fits) |>
  tidy(conf.int = TRUE)

mar_mi
```

. . .

-12.1 (95% CI -13.0, -11.1): the interval covers the benchmark, and the
standard error pays for the imputation instead of hiding it

## What the verb adds

- The effect labels survive: the pooled result still knows it is a marginal
  difference
- The complete-data degrees of freedom are read from the outcome models, with
  a small-sample adjustment
- `mice::pool()` also works with these results
- Results from balancing weights pool with the same verb

## {.center background-color="#23373B"}

### **What imputation cannot fix**

## Missingness from the value itself, again

```{r}
p_missing <- plogis(-0.5 + 3 * as.numeric(scale(net_data$malaria_risk)))
set.seed(789)
missing_mnar <- rbinom(n, 1, p_missing) == 1

net_data_mnar <- net_data |>
  select(net, income, health, temperature, malaria_risk) |>
  mutate(malaria_risk = if_else(missing_mnar, NA_real_, malaria_risk))
```

. . .

This is structure 3: the worse the malaria risk, the less likely we measured
it, and about 41% of the values are gone

## Complete cases cannot recover it

```{r}
#| output-location: fragment

mnar_complete_case <- fit_ipw_net(drop_na(net_data_mnar)) |>
  tidy(conf.int = TRUE)

mnar_complete_case
```

. . .

-9.4 (95% CI -9.9, -8.9): far from the benchmark, with an interval that
excludes it entirely

## The identical imputation

```{r}
imp_mnar <- net_data_mnar |>
  mutate(
    net_income = as.numeric(net) * income,
    net_health = as.numeric(net) * health,
    net_temperature = as.numeric(net) * temperature
  ) |>
  mice(m = 40, method = "pmm", seed = 20260814, print = FALSE)
```

. . .

Same frame, same method, same number of imputations, same seed

## The identical pipeline

```{r}
fits_mnar <- with(imp_mnar, {
  ps <- glm(net ~ income + health + temperature, family = binomial())
  w <- wt_ate(fitted(ps), net, exposure_type = "binary")
  om <- lm(malaria_risk ~ net, weights = w)
  ipw(ps, om)
})
```

## The identical pooling

```{r}
#| output-location: fragment

mnar_mi <- pool_ipw(fits_mnar) |>
  tidy(conf.int = TRUE)

mnar_mi
```

. . .

-10.7 (95% CI -11.3, -10.0): closer to the benchmark than the complete cases,
but the interval still excludes it

## No amount of modeling can rescue it

- When missingness depends on the missing values themselves, conditioning on
  the indicator opens a backdoor path that nothing observed can close
- The imputation model only sees the observed values, and under this mechanism
  they are a truncated sample
- Multiple imputation moves the estimate toward the benchmark, but it cannot
  reach it

. . .

Exactly what the DAG told us: there is no way to close the path

## When you suspect this structure

- Sensitivity analysis is how you probe what you cannot verify: the tipping
  point module's tools apply to missingness too
- When missingness is designed into a study as censoring, weight for it
  directly with inverse probability of censoring weights, covered in the
  selection bias module

## Every analysis in one picture

```{r}
#| echo: false

benchmark_result <- fit_ipw_net(net_data) |>
  tidy(conf.int = TRUE)

mar_label <- "Missing from net use and income (observed)"
mnar_label <- "Missing from malaria risk itself (unobserved)"

remedy_results <- bind_rows(
  benchmark_result |>
    mutate(analysis = "Full data", mechanism = "No missingness"),
  mar_complete_case |>
    mutate(analysis = "Complete case", mechanism = mar_label),
  mar_regression |>
    mutate(analysis = "Regression imputation", mechanism = mar_label),
  mar_mi |>
    mutate(analysis = "Multiple imputation", mechanism = mar_label),
  mnar_complete_case |>
    mutate(analysis = "Complete case", mechanism = mnar_label),
  mnar_mi |>
    mutate(analysis = "Multiple imputation", mechanism = mnar_label)
)

remedy_results |>
  mutate(
    analysis = fct_rev(fct_inorder(analysis)),
    mechanism = fct_inorder(mechanism)
  ) |>
  ggplot(aes(estimate, analysis, color = mechanism)) +
  geom_vline(
    xintercept = benchmark_result$estimate,
    linetype = "dashed",
    color = "grey60"
  ) +
  geom_point() +
  geom_segment(aes(x = conf.low, xend = conf.high, yend = analysis)) +
  facet_wrap(~mechanism, ncol = 1, scales = "free_y") +
  labs(x = "IPW effect of net use on malaria risk", y = NULL) +
  theme(legend.position = "none") +
  ggokabeito::scale_color_okabe_ito()
```

## What the picture says

- Under a recoverable structure, every remedy sits near the benchmark, and
  multiple imputation prices the uncertainty honestly
- Under missingness driven by the values themselves, nothing in that panel is
  trustworthy: both estimates miss, and their intervals are confidently wrong
- The data alone look similar in both worlds; the structure is what told us
  which world we were in

## {.center background-color="#23373B"}

### **The real thing is worse**

## Back to the real missingness

- In the malaria data we knew the answer; in the Disney data we never will
- The question: the effect of Extra Magic Mornings on the *actual* wait time
  at 9 AM
- The 6.2 minutes from the outcome model exercises was the effect on *posted*
  wait times, a different outcome

. . .

207 of 354 days are missing the actual wait time, and no benchmark is coming
to save us

## The pipeline for actual wait times {.small}

```{r}
fit_ipw_actual <- function(.data) {
  propensity_model <- glm(
    park_extra_magic_morning ~
      park_ticket_season + park_close + park_temperature_high,
    family = binomial(),
    data = .data
  )
  weighted_data <- propensity_model |>
    augment(type.predict = "response", data = .data) |>
    mutate(
      wts = wt_ate(.fitted, park_extra_magic_morning, exposure_type = "binary")
    )
  ipw_model <- lm(
    wait_minutes_actual_avg ~ park_extra_magic_morning,
    data = weighted_data,
    weights = wts
  )
  ipw(propensity_model, ipw_model)
}
```

## Complete case: 147 days

```{r}
#| output-location: fragment

seven_dwarfs_complete <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9, !is.na(wait_minutes_actual_avg))

fit_ipw_actual(seven_dwarfs_complete) |>
  tidy(conf.int = TRUE)
```

. . .

5.1 minutes (95% CI -10.8, 21.0): an interval four times wider than the
posted-wait analysis's, and it says almost nothing

## Impute the full data

```{r}
seven_dwarfs_to_impute <- seven_dwarfs_train_2018 |>
  mutate(park_ticket_season = as.factor(park_ticket_season)) |>
  select(
    park_date,
    wait_minutes_actual_avg,
    wait_minutes_posted_avg,
    wait_hour,
    park_temperature_high,
    park_close,
    park_ticket_season,
    park_extra_magic_morning
  )
```

. . .

- Every variable in the propensity score and outcome models, the outcome
  itself, and one strong auxiliary predictor: the posted wait time
- The frame also carries `wait_hour`, a structural auxiliary that lets the
  imputation borrow across hours, and `park_date`, which the predictor matrix
  will exclude

## Ten imputations, to start

```{r}
predictor_matrix <- make.predictorMatrix(seven_dwarfs_to_impute)
# park_date identifies rows, so it must not predict imputed values
predictor_matrix[, 1] <- 0

imp_disney <- mice(
  seven_dwarfs_to_impute,
  m = 10,
  predictorMatrix = predictor_matrix,
  method = "pmm",
  seed = 1,
  print = FALSE
)
```

. . .

Ten imputations against 88.7% missing: far below the rule of thumb

## Fit the pipeline in each completed dataset

```{r}
#| output-location: fragment

# mice masks tidyr::complete(), so name the package explicitly
disney_fits <- mice::complete(imp_disney, action = "all") |>
  map(\(.df) fit_ipw_actual(filter(.df, wait_hour == 9)))

disney_estimates <- map_dfr(disney_fits, tidy)

round(disney_estimates$estimate, 2)
```

. . .

Ten completed datasets, ten answers: from -2.3 to 11.5 minutes

## Pool the ten

```{r}
#| output-location: fragment

pool_ipw(disney_fits) |>
  tidy(conf.int = TRUE)
```

. . .

4.2 minutes (95% CI -10.6, 19.0): pooling worked, and the interval is still
almost as wide as the complete-case one

## The honest lessons

- Ten imputations against 88.7% incompleteness violates the rule of thumb;
  with more imputations, the pooled estimate varies far less from run to run
- No number of imputations makes the interval informative: 207 of the 354
  outcome values are simply gone
- Multiple imputation prices that honestly, where deterministic imputation
  would have hidden it

. . .

In the exercises, you will impute the analysis subset directly, with `m`
matched to the rule of thumb

## *Your Turn 1*

### Open **`15-bonus-missingness-exercises.qmd`**

### Count the days missing the actual wait time and visualize the missingness pattern

### Decide which missingness DAG you believe, and what it implies

`r countdown::countdown(minutes = 5)`

## *Your Turn 2*

### Impute the missing actual wait times with a regression model that keeps the exposure and confounders in their analysis forms

### Estimate the effect of an Extra Magic Morning on the actual wait time with the IPW pipeline

`r countdown::countdown(minutes = 8)`

## *Your Turn 3*

### Multiply impute the hour 9 data with `mice()`, choosing `m` by the rule of thumb

### Fit the full IPW pipeline within each imputation and pool with `pool_ipw()`

### Finish early? Try the *stretch goals*

`r countdown::countdown(minutes = 10)`
