---
title: "Bonus: Missing Data"
author: "Malcolm Barrett"
institute: "Stanford University"
format: "kakashi-revealjs"
---

## {background-color="#23373B"}

```{r}
#| label: setup
#| include: false

options(
  tibble.max_extra_cols = 6,
  tibble.width = 60
)
library(tidyverse)
library(broom)
library(propensity)
library(causalworkshop)
library(touringplans)
# mice masks tidyr::complete(), so tidyr::complete() calls must be namespaced
library(mice)
library(ggmice)
library(ggdag)
library(patchwork)

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
)
```

- `lm()` and `glm()` drop rows with missing values silently
- Every model we have fit in this course has been a complete-case analysis

## How much is missing?

```{r}
#| output-location: fragment

seven_dwarfs_train_2018 |>
  filter(wait_hour == 9) |>
  summarize(
    days = n(),
    missing_actual = sum(is.na(wait_minutes_actual_avg)),
    prop_missing = mean(is.na(wait_minutes_actual_avg))
  )
```

. . .

At the 9 AM hour, 207 of 354 days are missing the actual wait time; across the
full data, 88.7% of actual wait times are missing

## A known benchmark

- The malaria data has no missing values, so we will remove some on purpose
- Then we compare each remedy to the answer we already know

. . .

The benchmark: the full-data IPW estimate of the effect of net use on malaria
risk, -12.5 (from the outcome model module)

## {.center background-color="#23373B"}

### **The missingness indicator**

## Two nodes for one variable

- We draw two nodes for a given variable: one represents the true values, and
  the other represents a *missingness indicator*, whether we observed the value
  or not
- We are always conditioning on the data we actually have

## Two nodes for one variable

With missingness, that usually means conditioning on *complete* observations:
the subset where every variable we need is observed

## A DAG with missingness

```{r}
#| echo: false

malaria_labels <- c(
  malaria_risk = "Risk of malaria",
  malaria_missing = "Missingness\nin malaria risk",
  net = "Mosquito net",
  unknown = "Unknown\nmechanism"
)

missing_malaria <- dagify(
  malaria_risk ~ net,
  malaria_missing ~ net + unknown,
  coords = list(
    x = c(net = 0, unknown = 0, malaria_risk = 2, malaria_missing = 2),
    y = c(net = 1, unknown = -1, malaria_risk = 1, malaria_missing = -1)
  ),
  labels = malaria_labels,
  exposure = "net",
  outcome = "malaria_risk"
)

missing_malaria |>
  ggdag(use_text = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  expand_plot(expansion(0.2), expansion(0.2))
```

## {.center background-color="#23373B"}

### **When complete cases are enough**

## Conditioning on the indicator

```{r}
#| echo: false

missing_malaria |>
  ggdag_adjust("malaria_missing", use_text = FALSE, collider_lines = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot(expansion(0.2), expansion(0.2)) +
  ggokabeito::scale_color_okabe_ito()
```

## Recoverable effects

- The causal effect is *recoverable*: we can still estimate it without bias
  with complete-case analysis
- Because we only use complete observations, we have a smaller sample size and
  thus reduced precision

. . .

We are unable to correctly estimate the *mean* of malaria risk because values
are systematically missing by net use

## {.center background-color="#23373B"}

### **When complete cases are not enough**

## Missingness from the value itself

- Is it feasible that the missingness of malaria risk is related to its own
  values?
- It could be that a person's malaria risk is so severe that the measurement
  never happens

## An arrow from the outcome

```{r}
#| echo: false

missing_malaria_own <- dagify(
  malaria_risk ~ net,
  malaria_missing ~ net + unknown + malaria_risk,
  coords = list(
    x = c(net = 0, unknown = 0, malaria_risk = 2, malaria_missing = 2),
    y = c(net = 1, unknown = -1, malaria_risk = 1, malaria_missing = -1)
  ),
  labels = malaria_labels,
  exposure = "net",
  outcome = "malaria_risk"
)

missing_malaria_own |>
  ggdag(use_text = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  expand_plot(expansion(0.2), expansion(0.2))
```

## Conditioning now opens a backdoor

```{r}
#| echo: false

missing_malaria_own |>
  ggdag_adjust("malaria_missing", use_text = FALSE, collider_lines = FALSE) +
  geom_dag_text_repel(
    aes(label = label),
    color = "black",
    point.padding = 25,
    box.padding = 0.5,
    seed = 1
  ) +
  theme_dag() +
  theme(legend.position = "none") +
  expand_plot(expansion(0.2), expansion(0.2)) +
  ggokabeito::scale_color_okabe_ito()
```

## No way to close it

- Whether malaria risk was measured is now a descendant of both the exposure
  and the outcome
- Conditioning on the indicator opens a backdoor path, creating
  nonexchangeability

. . .

There is no way to close the backdoor path opened by conditioning on
missingness

## More than backdoor paths

- What effects we can recover is more complicated than determining backdoor
  paths alone

## More than backdoor paths

- When there are no backdoor paths, we can calculate the causal effect, but we
  may not be able to calculate other statistics, like the mean of the exposure
  or outcome
- When conditioning on missingness opens a backdoor path, sometimes we can
  close it, and sometimes we cannot

## {.center background-color="#23373B"}

### **Five structures of missingness**

## Five DAGs

- Consider five DAGs, where `a` is malaria risk, `n` is net use, `u` is
  unknown, and `m` is missing
- Each represents a simple but differing structure of missingness

. . .

In DAGs 1-3, malaria risk has missingness; in DAGs 4-5, net use does

## Five structures

```{r}
#| echo: false

define_dag <- function(..., tag, title) {
  dagify(
    ...,
    coords = time_ordered_coords(),
    exposure = "n",
    outcome = "a"
  ) |>
    ggdag(size = 0.7) +
    labs(title = paste0(tag, ": ", title)) +
    theme_dag() +
    theme(plot.title = element_text(size = 12)) +
    expand_plot(expansion(c(0.4, 0.4)), expansion(c(0.4, 0.4)))
}

dag_1 <- define_dag(
  a ~ n,
  m ~ u,
  tag = "1",
  title = "malaria risk is missing"
)

dag_2 <- define_dag(
  a ~ n,
  m ~ u + n,
  tag = "2",
  title = "malaria risk is missing"
)

dag_3 <- define_dag(
  a ~ n,
  m ~ u + a,
  tag = "3",
  title = "malaria risk is missing"
)

dag_4 <- define_dag(
  a ~ n,
  m ~ u + n,
  tag = "4",
  title = "net use is missing"
)

dag_5 <- define_dag(
  a ~ n,
  m ~ u + a,
  tag = "5",
  title = "net use is missing"
)

(dag_1 + dag_2 + dag_3) / (plot_spacer() + dag_4 + dag_5)
```

```{r}
#| echo: false

set.seed(123)
n <- nrow(net_data)
m1 <- as.logical(rbinom(n, 1, 0.35))
m2 <- as.logical(
  if_else(net_data$net, rbinom(n, 1, 0.9), rbinom(n, 1, 0.15))
)
# the threshold 38 sits near the median of malaria risk
m3 <- as.logical(
  if_else(net_data$malaria_risk > 38, rbinom(n, 1, 0.9), rbinom(n, 1, 0.05))
)
# the same structures, but net use gets the resulting missingness
m4 <- m2
m5 <- m3
```

## What we can estimate

```{r}
#| echo: false

# fit_ipw_net() is defined again on the visible pipeline slide below; keep
# the two definitions in sync
fit_ipw_net <- function(.data) {
  propensity_model <- glm(
    net ~ income + health + temperature,
    family = binomial(),
    data = .data
  )
  weighted_data <- propensity_model |>
    augment(type.predict = "response", data = .data) |>
    mutate(wts = wt_ate(.fitted, net, exposure_type = "binary"))
  ipw_model <- lm(malaria_risk ~ net, data = weighted_data, weights = wts)
  ipw(propensity_model, ipw_model)
}

fit_stats <- function(dag, miss, missing_for) {
  d <- net_data
  if (missing_for == "outcome") {
    d$malaria_risk[miss] <- NA
  } else {
    d$net[miss] <- NA
  }

  t_risk <- t.test(d$malaria_risk)
  p_net <- prop.test(sum(d$net, na.rm = TRUE), sum(!is.na(d$net)))
  # ipw() requires both models to see the same rows, so filter to complete
  # cases before fitting instead of letting lm() drop rows
  complete_cases <- d |> filter(!is.na(malaria_risk), !is.na(net))
  effect <- tidy(fit_ipw_net(complete_cases), conf.int = TRUE)

  tibble(
    dag = dag,
    prop_net_estimate = unname(p_net$estimate),
    prop_net_lower = p_net$conf.int[[1]],
    prop_net_upper = p_net$conf.int[[2]],
    mean_risk_estimate = unname(t_risk$estimate),
    mean_risk_lower = t_risk$conf.int[[1]],
    mean_risk_upper = t_risk$conf.int[[2]],
    effect_estimate = effect$estimate,
    effect_lower = effect$conf.low,
    effect_upper = effect$conf.high
  ) |>
    pivot_longer(
      cols = -dag,
      names_to = c("stat", ".value"),
      names_pattern = "^(.*)_(estimate|lower|upper)$"
    )
}

dag_stats <- bind_rows(
  fit_stats("No missingness", rep(FALSE, n), "outcome"),
  fit_stats("DAG 1", m1, "outcome"),
  fit_stats("DAG 2", m2, "outcome"),
  fit_stats("DAG 3", m3, "outcome"),
  fit_stats("DAG 4", m4, "exposure"),
  fit_stats("DAG 5", m5, "exposure")
)

dag_stats |>
  mutate(
    true_value = if_else(
      dag == "No missingness",
      "True value",
      "Observed value"
    ),
    dag = factor(dag, levels = c(paste("DAG", 5:1), "No missingness")),
    stat = factor(
      stat,
      levels = c("prop_net", "mean_risk", "effect"),
      labels = c(
        "Proportion using nets",
        "Mean of malaria risk",
        "Causal effect"
      )
    )
  ) |>
  ggplot(aes(color = true_value)) +
  geom_point(aes(estimate, dag)) +
  geom_segment(aes(
    x = lower,
    xend = upper,
    y = dag,
    yend = dag,
    group = stat
  )) +
  facet_wrap(~stat, scales = "free_x") +
  labs(y = NULL, color = NULL) +
  ggokabeito::scale_color_okabe_ito()
```

## Further reading

See Moreno-Betancur et al. (2018) for a comprehensive overview of what effects
are recoverable from different structures of missingness

::: tiny
Moreno-Betancur M, Lee KJ, Leacy FP, White IR, Simpson JA, Carlin JB (2018).
Canonical causal diagrams to guide the treatment of missing data in
epidemiologic studies. American Journal of Epidemiology.
doi:10.1093/aje/kwy173
:::

## {.center background-color="#23373B"}

### **MCAR, MAR, and MNAR**

## Three terms you will see

In the grand tradition of statisticians being bad at naming things, you will
also commonly see missingness discussed in terms of *missing completely at
random*, *missing at random*, and *missing not at random*

. . .

We can explain these ideas by the causal structure of missingness and the
availability of variables and values related to that structure in your data

## Missing completely at random

- **MCAR**: there are missing values, but the causes of missingness are such
  that the missingness process is unrelated to the causal structure of your
  question
- In other words, the only problem with missingness is a reduction in sample
  size

. . .

This is structure 1, where the only cause of missingness is the unknown
mechanism

## Missing at random

- **MAR**: the causes of missingness are related to the causal structure of
  the research problem, but missingness only depends on the variables and
  values in the data that we have actually observed

. . .

This is structure 2, where the missingness of malaria risk depends on net use,
which we always observe

## Missing not at random

- **MNAR**: the causes of missingness are related to the causal structure of
  the research problem, but the process is related to values we are missing
- A classic example is when a variable's missingness is impacted by itself:
  higher values of `x` are more likely to be missing in `x`, as in structure 3

. . .

We do not have that information because, by definition, it is missing

## {.center background-color="#23373B"}

### **Regression imputation**

## A missingness mechanism

```{r}
p_missing <- plogis(
  -0.8 + 1.4 * net_data$net + 0.3 * as.numeric(scale(net_data$income))
)
set.seed(88)
missing_mar <- rbinom(n, 1, p_missing) == 1

net_data_mar <- net_data |>
  select(net, income, health, temperature, malaria_risk) |>
  mutate(malaria_risk = if_else(missing_mar, NA_real_, malaria_risk))
```

. . .

Malaria risk now goes missing more often for net users and for higher-income
households: about 40% of the values are missing, for reasons we can see in
the data

## The IPW pipeline

```{r}
fit_ipw_net <- function(.data) {
  propensity_model <- glm(
    net ~ income + health + temperature,
    family = binomial(),
    data = .data
  )
  weighted_data <- propensity_model |>
    augment(type.predict = "response", data = .data) |>
    mutate(wts = wt_ate(.fitted, net, exposure_type = "binary"))
  ipw_model <- lm(malaria_risk ~ net, data = weighted_data, weights = wts)
  ipw(propensity_model, ipw_model)
}
```

. . .

The pipeline from the outcome model module: propensity score model, ATE
weights, weighted outcome model, `ipw()`

## Complete-case analysis

```{r}
#| output-location: fragment

net_data_complete <- net_data_mar |> drop_na()

mar_complete_case <- fit_ipw_net(net_data_complete) |>
  tidy(conf.int = TRUE)

mar_complete_case
```

. . .

-12.8 (95% CI -14.3, -11.2): the effect survives, with a wider interval, as
the missingness DAG suggested

## Complete-case analysis

- We filtered to the complete cases ourselves before fitting either model
- Had we let `lm()` drop the missing rows silently, `ipw()` would error: the
  propensity score model would see all 1752 rows and the outcome model only
  the 1047 complete ones

## Complete-case analysis

- Complete-case analysis is an analytic choice, and here we have to make it
  explicitly

. . .

The effect is recoverable here, but the mean of malaria risk is not: the
complete cases average 42.6 against the full-data 39.7

## Regression imputation

- The analysis variables predict malaria risk well, so we can use them to fill
  in the values we failed to measure
- Fit a model for malaria risk on the complete cases, then plug its prediction
  in wherever the value is missing

## Regression imputation

- When fitting imputation models, include every variable that appears in your
  later models, in the same form it takes there

## Regression imputation

```{r}
imputation_model <- lm(
  malaria_risk ~ net + income + health + temperature,
  data = net_data_mar
)

net_data_imputed <- imputation_model |>
  augment(newdata = net_data_mar) |>
  mutate(malaria_risk = coalesce(malaria_risk, .fitted))
```

. . .

`coalesce()` keeps the observed value where we have one and uses the model's
prediction where we do not

## The imputed estimate

```{r}
#| output-location: fragment

mar_regression <- fit_ipw_net(net_data_imputed) |>
  tidy(conf.int = TRUE)

mar_regression
```

. . .

-12.5 (95% CI -13.1, -11.8): very close to the benchmark

## Standard errors after imputation

- Each of the 705 missing values got a single predicted value, and the outcome
  model then treated it like real data

## Standard errors after imputation

- The standard error is 0.34, *narrower* than the 0.44 we had with no missing
  data at all
- Imputed values cannot make us more certain than measured ones

## Standard errors after imputation

- If you use this approach, include the fitting of the imputation model in
  your bootstrap so the standard errors reflect its uncertainty

. . .

Deterministic imputation understates uncertainty

## {.center background-color="#23373B"}

### **Multiple imputation**

## Single imputation is inefficient

- Regression imputation relies on a single imputed value for each missing
  observation
- Done correctly, with the imputation model inside the bootstrap, this plug-in
  approach is generally inefficient: we end up with wide confidence intervals,
  reflecting the fact that we relied on a single imputed value

## Single imputation is inefficient

Multiple imputation instead draws predicted values from a distribution,
capturing the uncertainty inherent in the missing data

## Many completed datasets

- We generate several completed datasets whose imputed values differ from draw
  to draw
- The values we observed are identical in every dataset; the values we imputed
  vary
- The variation between datasets is the uncertainty from the imputation

## Two rules for causal imputation models

1. The imputation model must include *every variable* in the outcome model and
   the propensity score model: the confounders, the exposure, and the outcome
   itself

## Two rules for causal imputation models

2. Mirror the same functional forms you will use downstream: if a variable
   gets a spline or an interaction later, it needs one in the imputation model
   too

## A typical causal analysis

1. Impute multiple completed datasets with a model that includes the outcome
   and all covariates
2. Estimate the treatment effect within each dataset: propensity score model,
   weights, weighted outcome model

## A typical causal analysis

3. Pool the estimates with Rubin's rules to get one effect and valid standard
   errors

. . .

Rule of thumb: the number of imputations should roughly match the percentage
of incomplete cases

## The missingness pattern

`plot_pattern()` shows which variables are missing and in what combinations:
here, one block of 705 rows missing malaria risk

```{r}
#| output-location: slide

library(ggmice)
plot_pattern(net_data_mar)
```

## Interactions in the imputation model

- In the malaria data, the effect of net use on malaria risk varies with
  income
- An imputation model with only main effects would fill in values from an
  effect that is too weak, and the pooled estimate would inherit that bias

## Interactions in the imputation model

- Rule 2 again: the marginal effect averages over that heterogeneity, so the
  imputation model has to represent it

## Build the imputation frame

```{r}
net_data_to_impute <- net_data_mar |>
  mutate(
    net_income = as.numeric(net) * income,
    net_health = as.numeric(net) * health,
    net_temperature = as.numeric(net) * temperature
  )
```

. . .

The product columns are complete, so `mice()` never imputes them; they serve
as predictors in the imputation model for malaria risk

## Impute with mice()

```{r}
library(mice)
imp <- mice(
  net_data_to_impute,
  m = 40,
  method = "pmm",
  seed = 20260814,
  print = FALSE
)
```

## Impute with mice()

- Predictive mean matching fills each missing value with an observed value
  borrowed from a case whose prediction is close
- About 40% of the cases are incomplete, so the rule of thumb says about 40
  imputations; we use `m = 40`

## Observed versus imputed

The imputed values should be plausible, and their distribution should reflect
the mechanism

```{r}
#| output-location: slide

ggmice(imp, aes(x = malaria_risk, group = .imp)) +
  geom_density()
```

## Reading the densities

- The observed values (blue) center higher than the imputed values (red)
- That is what the mechanism implies: net users are missing malaria risk more
  often, and net users have lower risk

## Reading the densities

- The imputation model can see this because net use is one of its predictors

## Forty completed datasets

```{r}
#| output-location: fragment

# mice masks tidyr::complete(), so name the package explicitly
completed <- mice::complete(imp, action = "all")

length(completed)
```

. . .

Each element is the whole dataset with the missing malaria risk values filled in

## Fit within each imputation

```{r}
fits <- completed |>
  map(\(.df) {
    ps <- glm(
      net ~ income + health + temperature,
      family = binomial(),
      data = .df
    )
    w <- wt_ate(fitted(ps), .df$net, exposure_type = "binary")
    om <- lm(malaria_risk ~ net, data = .df, weights = w)
    ipw(ps, om)
  })
```

## Fit within each imputation

`map()` runs the pipeline once per completed dataset: forty propensity score
models, forty sets of weights, forty outcome models, forty `ipw()` results

## Refit the propensity model every time

- The whole analysis is rebuilt inside each imputation, including the
  propensity score model
- Averaging the propensity scores across imputations and weighting with one
  averaged set hides the propensity model's estimation uncertainty from the
  pooling step

## Refit the propensity model every time

- Here only the outcome is imputed, so each refit returns the same model; when
  confounders or the exposure have missingness, the refits differ, and that
  variation belongs in the pooled standard error

. . .

The propensity score model is part of the analysis, so it belongs inside the
imputation loop

## Rubin's rules

```{r}
#| output-location: fragment

estimates <- fits |>
  map(tidy) |>
  list_rbind()

estimates |>
  summarize(
    pooled = mean(estimate),
    within_var = mean(std.error^2),
    between_var = var(estimate),
    pooled_se = sqrt(within_var + (1 + 1 / n()) * between_var)
  )
```

. . .

The pooled estimate is the average across imputations; its variance is the
average within-imputation variance plus the between-imputation variance,
inflated by `1 + 1/m`

## Pooling with `pool_ipw()`

```{r}
#| output-location: fragment

mar_mi <- pool_ipw(fits) |>
  tidy(conf.int = TRUE)

mar_mi
```

. . .

-12.1 (95% CI -13.0, -11.1): the interval covers the benchmark, and the
standard error accounts for the imputation

## {.center background-color="#23373B"}

### **What imputation cannot fix**

## Missingness from the value itself

```{r}
p_missing <- plogis(-0.5 + 3 * as.numeric(scale(net_data$malaria_risk)))
set.seed(789)
missing_mnar <- rbinom(n, 1, p_missing) == 1

net_data_mnar <- net_data |>
  select(net, income, health, temperature, malaria_risk) |>
  mutate(malaria_risk = if_else(missing_mnar, NA_real_, malaria_risk))
```

. . .

This is structure 3: the worse the malaria risk, the less likely we measured
it, and about 41% of the values are gone

## Complete-case analysis

```{r}
#| output-location: fragment

mnar_complete_case <- fit_ipw_net(drop_na(net_data_mnar)) |>
  tidy(conf.int = TRUE)

mnar_complete_case
```

. . .

-9.4 (95% CI -9.9, -8.9): far from the benchmark, with an interval that
excludes it entirely

## The same analysis

```{r}
imp_mnar <- net_data_mnar |>
  mutate(
    net_income = as.numeric(net) * income,
    net_health = as.numeric(net) * health,
    net_temperature = as.numeric(net) * temperature
  ) |>
  mice(m = 40, method = "pmm", seed = 20260814, print = FALSE)
```

. . .

Same frame, same method, same number of imputations, same seed

## The same analysis

```{r}
fits_mnar <- with(imp_mnar, {
  ps <- glm(net ~ income + health + temperature, family = binomial())
  w <- wt_ate(fitted(ps), net, exposure_type = "binary")
  om <- lm(malaria_risk ~ net, weights = w)
  ipw(ps, om)
})
```

## The same analysis

```{r}
#| output-location: fragment

mnar_mi <- pool_ipw(fits_mnar) |>
  tidy(conf.int = TRUE)

mnar_mi
```

. . .

-10.7 (95% CI -11.3, -10.0): closer to the benchmark than the complete cases,
but the interval still excludes it

## Why imputation cannot help

- When missingness depends on the missing values themselves, conditioning on
  the indicator opens a backdoor path that nothing observed can close
- The imputation model only sees the observed values, and under this mechanism
  they are a truncated sample

## Why imputation cannot help

- Multiple imputation moves the estimate toward the benchmark, but it does not
  reach it

. . .

The DAG said as much: there is no way to close the path

## What to do instead

- Sensitivity analyses let you probe what you cannot verify; the tools from
  the tipping point module apply to missingness as well
- When missingness is designed into a study as censoring, weight for it
  directly with inverse probability of censoring weights, covered in the
  selection bias module

## Comparing the estimates

```{r}
#| echo: false

benchmark_result <- fit_ipw_net(net_data) |>
  tidy(conf.int = TRUE)

mar_label <- "Missing from net use and income (observed)"
mnar_label <- "Missing from malaria risk itself (unobserved)"

remedy_results <- bind_rows(
  benchmark_result |>
    mutate(analysis = "Full data", mechanism = "No missingness"),
  mar_complete_case |>
    mutate(analysis = "Complete case", mechanism = mar_label),
  mar_regression |>
    mutate(analysis = "Regression imputation", mechanism = mar_label),
  mar_mi |>
    mutate(analysis = "Multiple imputation", mechanism = mar_label),
  mnar_complete_case |>
    mutate(analysis = "Complete case", mechanism = mnar_label),
  mnar_mi |>
    mutate(analysis = "Multiple imputation", mechanism = mnar_label)
)

remedy_results |>
  mutate(
    analysis = fct_rev(fct_inorder(analysis)),
    mechanism = fct_inorder(mechanism)
  ) |>
  ggplot(aes(estimate, analysis, color = mechanism)) +
  geom_vline(
    xintercept = benchmark_result$estimate,
    linetype = "dashed",
    color = "grey60"
  ) +
  geom_point() +
  geom_segment(aes(x = conf.low, xend = conf.high, yend = analysis)) +
  facet_wrap(~mechanism, ncol = 1, scales = "free_y") +
  labs(x = "IPW effect of net use on malaria risk", y = NULL) +
  theme(legend.position = "none") +
  ggokabeito::scale_color_okabe_ito()
```

## Comparing the estimates

- Under a recoverable structure, every remedy sits near the benchmark, and
  multiple imputation reflects the added uncertainty in its standard error
- Under missingness driven by the values themselves, neither estimate is
  trustworthy: both miss the benchmark, and neither interval includes it

## Comparing the estimates

- The data look much the same under both mechanisms; only the structure tells
  them apart

## *Your Turn*

### Work through Your Turns 1-3 in `15-bonus-missingness-exercises.qmd`

`r countdown::countdown(minutes = 20)`
