---
title: "Causal Inference with `group_by()` and `summarize()`"
format: html
---

```{r}
#| label: setup
library(tidyverse)
set.seed(1)
```

## Your Turn 1

Run this code to generate the simulated data set

```{r}
n <- 1000 
sim <- tibble(
  confounder = rbinom(n, 1, 0.5),
  p_exposure = case_when(
    confounder == 1 ~ 0.75,
    confounder == 0 ~ 0.25
  ),
  exposure = rbinom(n, 1, p_exposure),
  outcome = confounder + rnorm(n)
)
```

1. Group the dataset by `confounder` and `exposure`
2. Calculate the mean of the `outcome` for the groups

```{r}
sim |>
  group_by(______, ______) |>
  summarise(avg_y = mean(______)) |>
  # pivot the data so we can get the difference
  # between the exposure groups
  pivot_wider(
    names_from = exposure,
    values_from = avg_y,  
    names_prefix = "x_"
  ) |>
  summarise(estimate = x_1 - x_0) |>
  summarise(estimate = mean(estimate)) # note, we would need to weight this if the confounder groups were not equal sized
```

## Your Turn 2

Run the following code to generate `sim2`

```{r}
n <- 1000
sim2 <- tibble(
  confounder_1 = rbinom(n, 1, 0.5),
  confounder_2 = rbinom(n, 1, 0.5), 
  
  p_exposure = case_when(
    confounder_1 == 1 & confounder_2 == 1 ~ 0.75,
    confounder_1 == 0 & confounder_2 == 1 ~ 0.9,
    confounder_1 == 1 & confounder_2 == 0 ~ 0.2,
    confounder_1 == 0 & confounder_2 == 0 ~ 0.1,
  ),
  exposure = rbinom(n, 1, p_exposure),
  outcome = confounder_1 + confounder_2 + rnorm(n) 
)
```

1. Group the dataset by the confounders and exposure
2. Calculate the mean of the outcome for the groups

```{r}
sim2 |>
  group_by(_____, _____, _____) |>
  summarise(avg_y = mean(_____)) |>
  pivot_wider(names_from = exposure,
              values_from = avg_y,  
              names_prefix = "x_") |>
  summarise(estimate = x_1 - x_0, .groups = "drop") |>
  summarise(estimate = mean(estimate))  
```

## Your Turn 3

Run the following code to generate `sim3`

```{r}
n <- 10000 
sim3 <- tibble(
  confounder = rnorm(n), 
  p_exposure = exp(confounder) / (1 + exp(confounder)),
  exposure = rbinom(n, 1, p_exposure),
  outcome = confounder + rnorm(n) 
)
```

1. Use `ntile()` from dplyr to calculate a binned version of `confounder` called `confounder_q`. We'll create a variable with 5 bins.
2. Group the dataset by the binned variable you just created and exposure
3. Calculate the mean of the outcome for the groups

```{r}
sim3 |>
  mutate(confounder_q = _____(_____, 5)) |>
  group_by(_____, _____) |>
  summarise(avg_y = mean(_____)) |>
  pivot_wider(
    names_from = exposure,
    values_from = avg_y,  
    names_prefix = "x_"
  ) |>
  summarise(estimate = x_1 - x_0)  |>
  summarise(estimate = mean(estimate))  
```

# Take aways

* Sometimes correlation *is* causation!
* In simple cases, grouping by confounding variables can get us the right answer without a statistical model
* Propensity scores generalize the idea of summarizing exposure effects to any number of confounders. Although we'll use models for this process, the foundations are the same.
