---
title: "Propensity Scores"
output: html_document
---


```{r}
library(tidyverse)
library(broom)
library(causaldata)
library(ggdag)
```

Using the National Health and Nutrition Examination Survey Data (`nhefs_complete`), we are interested in the relationship between the **exposure**, `qsmk`: whether the participant quit smoking, and the **outcome**, `wt82_71`: their weight change in kilograms.

Below is a proposed DAG of the relationship between 4 confounders: `sex`, `age`, `smokeyrs` and `wt71` and the exposure and outcome. 

_Knit this document to see the DAG or refer to the slides_.

```{r}
set.seed(1234)
# set up DAG
smk_wt_dag <- dagify(
  # specify causes of quitting smoking and weight gain:
  qsmk ~ sex + age + smokeyrs + wt71,
  wt82_71 ~ sex + age + smokeyrs + wt71,
  # specify causal question:
  exposure = "qsmk", 
  outcome = "wt82_71",
  # set up labels:
  # here, I'll use the same variable names as the data set, but I'll label them
  # with clearer names
  labels = c(
    # causal question
    "qsmk" = "quit\nsmoking (qsmk)",
    "wt82_71" = "change in\nweight",
    
    # demographics
    "age" = "age",
    "sex" = "sex",
    
    # health
    "wt71" = "baseline\nweight (wt71)",
    
    # smoking history
    "smokeyrs" = "yrs of\nsmoking (smokeyrs)"
  )
) %>% 
  tidy_dagitty()

smk_wt_dag %>% 
  ggdag(text = FALSE, use_labels = "label") 
```

## Your Turn

_After updating the code chunks below, change `eval = TRUE` before knitting._

Fit a propensity score model for `qsmk` using the above proposed confounders.

```{r, eval = FALSE}
propensity_model <- ___(
  ___ ~ ___,
  data = nhefs_complete,
  family = _____
)
```

Add the propensity scores to the `nhefs_complete` data set, call this new dataset `df`.

```{r, eval = FALSE}
df <- propensity_model %>%
  ____(type.predict = ____, data = ____)
```


Stretch Goal 1: 

Examine two histograms of the propensity scores, one for those that quit smoking (`qsmk == 1`) and one for those that did not (`qsmk == 0`). How do these compare?

```{r}

```

