---
title: "Propensity Score Weighting"
output: html_document
---


```{r}
library(tidyverse)
library(broom)
library(causaldata)
```

Using the National Health and Nutrition Examination Survey Data (`nhefs_complete`), we are interested in the relationship between the **exposure**, `qsmk`: whether the participant quit smoking, and the **outcome**, `wt82_71`: their weight change in kilograms.

Below is the propensity score model you created in the previous exercise.

```{r, eval = FALSE}
propensity_model <- glm(
  qsmk ~ age + sex + wt71 + smokeyrs,
  data = nhefs_complete,
  family = binomial()
)

df <- propensity_model %>%
  augment(type.predict = "response", data = nhefs_complete)
```

## Your Turn (Matching)

_After updating the code chunks below, change `eval = TRUE` before knitting._

Create at "matched" data set using the same propensity score model as above and a caliper of 0.2.

```{r, eval = FALSE}
library(MatchIt)
m <- matchit(___,
        data = nhefs_complete,
        ___,
        ___
        )
matched_df <- ___(m)
```



## Your Turn (Weighting)

_After updating the code chunks below, change `eval = TRUE` before knitting._

Add the ATE weights to the data frame, `df`

```{r, eval = FALSE}
df <- df %>%
  mutate(w_ate = ___)
```


Stretch Goal 1: 

Add ATM weights to the data frame, `df`

```{r, eval = FALSE}
df <- df %>%
  mutate(w_atm = ___)
```

Stretch Goal 2: 

Update the code below to examine the distribution of the weighted sample. **HINT** the part that needs to be updated is the `weight` parameter in two of the `geom_histogram()` calls.

```{r, eval = FALSE}
d <- df %>%
  tidyr::spread(qsmk, .fitted, sep = "_p")
```


```{r, eval = FALSE}
ggplot(d) +
  geom_histogram(bins = 50, aes(qsmk_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(qsmk_p1, weight = ____), fill = "green", alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = qsmk_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = qsmk_p0, weight = ____, y = -..count..), fill = "blue", alpha = 0.5) + 
  ylab("count") + xlab("p") +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs) +
  theme_minimal() + 
  geom_rect(aes(xmin = 0.95, xmax = 1, ymin = 5, ymax = 100), fill = "#5DB854") + 
  geom_text(aes(x = 0.975, y = 50), label = "trt", angle = 270, color = "white") + 
  geom_rect(aes(xmin = 0.95, xmax = 1, ymin = -100, ymax = -5), fill = "#5154B8") + 
  geom_text(aes(x = 0.975, y = -50), label = "control", angle = 270, color = "white") 
