---
title: "Propensity Score Diagnostics"
output: html_document
---


```{r}
library(tidyverse)
library(survey)
library(broom)
library(causaldata)
```

Using the National Health and Nutrition Examination Survey Data (`nhefs_complete`), we are interested in the relationship between the **exposure**, `qsmk`: whether the participant quit smoking, and the **outcome**, `wt82_71`: their weight change in kilograms.

Below is the propensity score model and weights you created in the previous exercise.

```{r, eval = FALSE}
propensity_model <- glm(
  qsmk ~ age + sex + wt71 + smokeyrs,
  data = nhefs_complete,
  family = binomial()
)

df <- propensity_model %>%
  augment(type.predict = "response", data = nhefs_complete) %>%
  mutate(w_ate = 1 / ifelse(qsmk == 0, 1 - .fitted, .fitted))
```

## Your Turn 1

_After updating the code chunks below, change `eval = TRUE` before knitting._

Create the survey design object to incorporate the weights.

```{r, eval = FALSE}
svy_des <- ____(
  ids = ~ 1,
  data = ___,
  weights = ___
)
```

Create the **unweighted** standardized mean differences data frame

```{r, eval = FALSE}
smd_table_unweighted <- ____(
  vars = _____,
  strata = _____,
  data = ____,
  test = FALSE)
```

Create the **weighted** standardized mean differences data frame

```{r, eval = FALSE}
smd_table <- ____(
  vars = _____,
  strata = _____,
  data = ____,
  test = FALSE)
```

Create a data frame that merges `smd_table_unweighted` and `smd_table` and pivots it to prepare for plotting

```{r, eval = FALSE}
plot_df <- data.frame(
  var = rownames(____),                        
  Unadjusted = _____,                      
  Weighted = _____) %>%
  pivot_longer(-var, names_to = "Method", values_to = "SMD")
```

Create the Love Plot using ggplot

```{r, eval = FALSE}
ggplot(data = _____, 
       mapping = aes(x = ____, y = ____, group = ____, color = ____)) +
  geom_line() +
  geom_point() + 
  geom_hline(yintercept = 0.1, color = "black", size = 0.1) +  
  coord_flip()
```



## Your Turn 2

Create an unweighted ECDF for `smokeyrs` by those who quit smoking and those who did not.

```{r, eval = FALSE}
ggplot(df, aes(x = ____, group = ____, color = factor(____))) +
  ____() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab(____) + 
  ylab("Proportion <= x") 
```


Create an weighted ECDF for `smokeyrs` by those who quit smoking and those who did not.

```{r, eval = FALSE}
ecdf_1 <- df %>%
  filter(____) %>%
  arrange(____) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(____) %>%
  arrange(____) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = _____, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = ____, y = cum_pct), color = "#5154B8") + 
  xlab(____) + 
  ylab("Proportion <= x") 
```

