---
title: "Propensity Score Diagnostics"
output: html_document
---


```{r}
library(tidyverse)
library(broom)
library(smd)
library(causaldata)
```

Using the National Health and Nutrition Examination Survey Data (`nhefs_complete`), we are interested in the relationship between the **exposure**, `qsmk`: whether the participant quit smoking, and the **outcome**, `wt82_71`: their weight change in kilograms.

Below is the propensity score model and weights you created in the previous exercise.

```{r, eval = FALSE}
propensity_model <- glm(
  qsmk ~ age + sex + wt71 + smokeyrs,
  data = nhefs_complete,
  family = binomial()
)

df <- propensity_model %>%
  augment(type.predict = "response", data = nhefs_complete) %>%
  mutate(w_ate = 1 / ifelse(qsmk == 0, 1 - .fitted, .fitted))
```

## Your Turn 1

_After updating the code chunks below, change `eval = TRUE` before knitting._

Calculate the standardized mean differences with and without weights

```{r, eval = FALSE}
smds <- df %>% 
  summarise(
    across(
      # variables to calculate SMD for
      ________, 
      list(
        unweighted = ~_____(.x, qsmk)$estimate, 
        weighted = ~_____(.x, qsmk, w = ___)$estimate
      )
    )
  )
```

Pivot `smds` so that it is in tidy format with the columns "SMD", "variable", and "Method".

```{r, eval = FALSE}
plot_df <- smds %>% 
  pivot_longer(
    _______, 
    values_to = "___", 
    names_to = c("___", "___"),
    
    # don't change this one. We really mean _!
    names_sep = "_"
  )
```

Create the Love Plot using ggplot

```{r, eval = FALSE}
ggplot(
  data = ____,
  aes(x = abs(____), y = ____, group = ____, color = ____)
) +
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, color = "black", size = 0.1)
```



## Your Turn 2

Create an unweighted ECDF for `smokeyrs` by those who quit smoking and those who did not.

```{r, eval = FALSE}
ggplot(df, aes(x = ____, group = ____, color = factor(____))) +
  ____() +
  scale_color_manual("Quit smoking", values = c("#5154B8", "#5DB854"),
                     labels = c("Yes", "No")) + 
  xlab(____) + 
  ylab("Proportion <= x") 
```


Create an weighted ECDF for `smokeyrs` by those who quit smoking and those who did not.

```{r, eval = FALSE}
ecdf_1 <- df %>%
  filter(____) %>%
  arrange(____) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ecdf_0 <- df %>%
  filter(____) %>%
  arrange(____) %>%
  mutate(cum_pct = cumsum(w_ate) / sum(w_ate))

ggplot(ecdf_1, aes(x = _____, y = cum_pct)) +
  geom_line(color = "#5DB854") +
  geom_line(data = ecdf_0, aes(x = ____, y = cum_pct), color = "#5154B8") + 
  xlab(____) + 
  ylab("Proportion <= x") 
```

