---
title: "Causal inference is not just a statistics problem"
format: html
---

```{r}
#| label: setup
library(tidyverse)
library(quartets)
```

## Your turn 1

For each of the following 4 datasets, create a scatterplot looking at the relationship between `exposure` and `outcome`: 

* `causal_collider`
* `causal_confounding`
* `causal_mediator`
* `causal_m_bias`

(Alternatively, you can work with `causal_quartet`, which has all four datasets stacked on top of one another.)

```{r}

```

For each of the above 4 datasets, look at the correlation between `exposure` and `covariate`


```{r}

```

For each of the above 4 datasets, fit a linear model to examine the relationship between the `exposure` and the `outcome`

```{r}

```

## Your turn 2

For each of the following 4 datasets, fit a linear model examining the relationship between `outcome_followup` and `exposure_baseline` adjusting for `covariate_baseline`: 

* `causal_collider_time`
* `causal_confounding_time`
* `causal_mediator_time`
* `causal_m_bias_time`

(Alternatively, you can work with `causal_quartet_time`, which has all four datasets stacked on top of one another.)

```{r}

```

