---
title: "Outcome Model"
output: html_document
---


```{r}
library(tidyverse)
library(broom)
library(causaldata)
library(rsample)
```

Using the National Health and Nutrition Examination Survey Data (`nhefs_complete`), we are interested in the relationship between the **exposure**, `qsmk`: whether the participant quit smoking, and the **outcome**, `wt82_71`: their weight change in kilograms.

## Your turn

_After updating the code chunks below, change `eval = TRUE` before knitting._

Create a function called `ipw_fit` that fits the propensity score model from Exercise 03, incorporates the ATE weights calculated in Exercise 04, and fits a weighted outcome model.

```{r, eval = FALSE}
it_ipw <- function(split, ...) { 
  .df <-____
  
  # fit propensity score model
  
  # calculate ATE weights

  # fit correctly bootsrapped ipw model 
  lm(___ ~ ___, data = .df, weights = ___) %>% 
    tidy() 
}
```

Bootstrap this result 1000 times.

```{r, eval = FALSE}
ipw_results <- ____(___, 1000, apparent = TRUE) %>% 
  mutate(results = map(splits, _____)) 
```


Calculate the confidence interval

```{r, eval = FALSE}
boot_estimate <- ____(____, ____) %>% 
  filter(term == ____)
```


Stretch goal: Do the same for a model using matching.
