---
title: "Causal Assumptions"
format: html
---

This exercise is a discussion, so there is no code to run.
We'll work through the causal assumptions for a question presented in *Causal Inference: What if?*: does quitting smoking make you gain weight?

Causal modeling has a special place in the history of smoking research: the studies that demonstrated that smoking causes lung cancer were observational.
Thanks to other studies, we also know that, if you're already a smoker, quitting smoking reduces your risk of lung cancer.
However, some have observed that former smokers tend to gain weight.
Is this the result of quitting smoking, or does something else explain this effect?

The data that *Causal Inference: What if?* uses to analyze this question come from NHEFS, the National Health and Nutrition Examination Survey Epidemiologic Follow-up Study.
NHEFS is a longitudinal, observational study.
Participants were surveyed and examined starting in 1971 and followed up through 1982.
Some quit smoking between the two visits and some did not, and we can compare how much weight each group gained over that period.

Nothing about that comparison is causal on its own.
The three assumptions below are what would license reading it as the effect of quitting smoking on weight gain.

## Your turn

Discuss each assumption with your neighbors.
For each one, decide how plausible it is for this question and what you would do if you thought it was violated.

### Exchangeability

Exchangeability requires that the two groups have the same potential outcomes on average: whether someone quit smoking is independent of the weight they would gain under either choice.

- Who quits smoking, and why?
- What would make people who quit and people who kept smoking non-exchangeable?
- Which confounders would you need to measure to make the groups conditionally exchangeable, and which of those are hard to measure well?

This DAG encodes one set of answers to those questions.

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(ggdag)
library(ggokabeito)
```

```{r}
#| echo: false
#| fig-width: 9

smk_wt_dag <- dagify(
  qsmk ~ sex +
    race +
    age +
    education +
    smokeintensity +
    smokeyrs +
    exercise +
    active +
    wt71,
  wt82_71 ~ qsmk +
    sex +
    race +
    age +
    education +
    smokeintensity +
    smokeyrs +
    exercise +
    active +
    wt71,
  exposure = "qsmk",
  outcome = "wt82_71",
  coords = time_ordered_coords(),
  labels = c(
    "qsmk" = "quit\nsmoking",
    "wt82_71" = "change in\nweight",
    "age" = "age",
    "sex" = "sex",
    "race" = "race",
    "education" = "education",
    "wt71" = "baseline\nweight",
    "active" = "daily\nactivity\nlevel",
    "exercise" = "exercise",
    "smokeintensity" = "smoking\nintensity",
    "smokeyrs" = "yrs of\nsmoking"
  )
) |>
  tidy_dagitty() |>
  node_status()

smk_wt_dag |>
  ggplot(aes(x, y, xend = xend, yend = yend, color = status)) +
  geom_dag_edges() +
  geom_dag_point() +
  geom_dag_label_repel(
    aes(x, y, label = label),
    # the confounders share one column, so their labels need room to the left
    nudge_x = ifelse(smk_wt_dag$x == 1, -0.6, 0.6),
    box.padding = 1.25,
    inherit.aes = FALSE,
    max.overlaps = Inf,
    seed = 10,
    label.size = NA,
    label.padding = 0.1,
    size = 14 / 3
  ) +
  theme_dag(base_size = 14) +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off") +
  scale_color_okabe_ito(na.value = "grey90") +
  expand_plot(
    expand_x = expansion(c(0.45, 0.2)),
    expand_y = expansion(c(0.1, 0.1))
  )
```

### Positivity

Positivity requires that every participant has a non-zero probability of quitting and a non-zero probability of continuing.

- Is there anyone in this study who could never have quit, or who was certain to quit?
- Are there combinations of the confounders you just listed where quitting is impossible or guaranteed?
- Would such a violation be structural or stochastic, and what would you do about each?

### Consistency

Consistency requires that the potential outcome for quitting is the weight change we actually observe among those who quit.

- What exactly does "quitting smoking" mean? Are there different ways of quitting, and do they relate to the potential outcome of weight change?
- Could one participant quitting affect another participant's weight? Is interference plausible here?
