---
title: "Outcome Model"
format: html
---


```{r}
#| label: setup
library(tidyverse)
library(broom)
library(touringplans)
library(rsample)
library(propensity)

seven_dwarfs <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9)
```

We are interested in examining the relationship between whether there were "Extra Magic Hours" in the morning (the **exposure**) and the average wait time for the Seven Dwarfs Mine Train the same day between 9am and 10am (the **outcome**).

## Your turn

*After updating the code chunks below, change `eval: true` before rendering*

Create a function called `ipw_fit` that fits the propensity score model from Exercise 03, incorporates the ATE weights calculated in Exercise 04, and fits a weighted outcome model.

```{r}
#| eval: false
fit_ipw <- function(split, ...) { 
  .df <- ____
  
  # fit propensity score model
  
  # calculate ATE weights

  # fit correctly bootstrapped ipw model 
  lm(___ ~ ___, data = .df, weights = ___) |>
    tidy() 
}
```

Bootstrap this result 1000 times.

```{r}
#| eval: false
set.seed(1234)

ipw_results <- ____(___, 1000, apparent = TRUE) |>
  mutate(boot_fits = map(splits, _____)) 
```

Check out the distribution of estimates (**no need to change this code**)

```{r}
#| eval: false
ipw_results |>
  mutate(
    estimate = map_dbl(
      boot_fits,
      # pull the `estimate` for `park_extra_magic_morning` for each fit
      \(.fit) .fit |>
        filter(term == "park_extra_magic_morning") |>
        pull(estimate)
    )
  ) |>
  ggplot(aes(estimate)) +
  geom_histogram(fill = "#D55E00FF", color = "white", alpha = 0.8) + 
  theme_minimal()
```

Calculate the confidence interval

```{r}
#| eval: false
boot_estimate <- ____(____, ____) |>
  filter(term == ____)

boot_estimate
```


Stretch goal: Do the same for a model using matching.
